---
title: "Causal Modeling in R: Whole Game"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(rsample)
library(ggdag)
library(ggokabeito)
library(causalworkshop)
library(propensity)
library(halfmoon)
library(estimatr)
library(tipr)

set.seed(1234)
```

1. Specify causal question (e.g. target trial)
2. Draw assumptions (causal diagram)
3. Model assumptions (e.g., propensity)
4. Diagnose model (e.g., balance)
5. Estimate causal effects (e.g., IPW)
6. Sensitivity analysis

## {.center background-color="#23373B"}

### **We'll focus on the broader ideas behind each step and what they look like all together; we don't expect you to fully digest each idea. We'll spend the rest of the workshop taking up each step in detail**

## {.center background-color="#23373B"}

### **Does mosquito bed net use reduce malaria risk?**

## {.small}

```{r}
library(causalworkshop)
net_data
```

::: small
Thanks to Andrew Heiss for the data!
:::

## Do net users have lower malaria risk?

```{r}
#| echo: false

net_data |>
  ggplot(aes(malaria_risk, fill = net)) +
  geom_density(color = NA, alpha = 0.8) +
  scale_fill_okabe_ito(order = c(1, 5)) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "bottom") +
  labs(
    x = "malaria risk (0-100)",
    fill = "used a net"
  )
```

## Do net users have lower malaria risk?

```{r}
#| highlight-output: [4, 5]

# ~16 points lower risk for net users vs. non-users
net_data |>
  group_by(net) |>
  summarize(
    mean_malaria_risk = mean(malaria_risk),
    sd = sd(malaria_risk),
    .groups = "drop"
  )
```

## {.center background-color="#23373B"}

### **draw your assumptions**

##

```{r}
#| echo: false
#| fig-width: 7.5

mosquito_dag <- dagify(
  malaria_risk ~ net + income + health + temperature + resistance,
  net ~ income + health + temperature + eligible + household,
  eligible ~ income + household,
  health ~ income,
  # specify causal question:
  exposure = "net",
  outcome = "malaria_risk",
  # lay the nodes out by hand:
  coords = list(
    x = c(
      malaria_risk = 7,
      net = 3,
      income = 4,
      health = 5,
      temperature = 6,
      resistance = 8.5,
      eligible = 2,
      household = 1
    ),
    y = c(
      malaria_risk = 2,
      net = 2,
      income = 3,
      health = 1,
      temperature = 3,
      resistance = 2,
      eligible = 3,
      household = 2
    )
  ),
  # set up labels:
  labels = c(
    malaria_risk = "Risk of malaria",
    net = "Mosquito net",
    income = "Income",
    health = "Health",
    temperature = "Nighttime temperatures",
    resistance = "Insecticide resistance",
    eligible = "Eligible for program",
    household = "Number in the household"
  )
) |>
  tidy_dagitty()

plot_mosquito_dag <- function(.dag) {
  .dag |>
    node_status() |>
    ggplot(
      aes(x, y, xend = xend, yend = yend, color = status)
    ) +
    geom_dag_edges() +
    geom_dag_point() +
    geom_dag_label_repel(
      aes(x, y, label = label),
      box.padding = 3.5,
      inherit.aes = FALSE,
      max.overlaps = Inf,
      seed = 10,
      label.size = NA,
      label.padding = 0.1,
      size = 14 / 3
    ) +
    xlim(0.5, 9) +
    ylim(0.5, 3.5) +
    theme_dag(base_size = 14) +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    scale_color_okabe_ito(na.value = "grey90")
}

mosquito_dag |>
  filter(name %in% c("net", "malaria_risk")) |>
  plot_mosquito_dag()
```

##

```{r}
#| echo: false
#| fig-width: 7.5

plot_mosquito_dag(mosquito_dag)
```

## {.center background-color="#23373B"}

### What do I need to control for?

## One causal path, seven confounding

```{r}
#| echo: false
#| fig-width: 15

glyph <- function(data, params, size) {
    data$shape <- 15
    data$size <- 12
    ggplot2::draw_key_point(data, params, size)
}

mosquito_dag |>
    dag_paths() |>
    mutate(
        effects = case_when(
            set == "1" & path == "open path" ~ "true effect",
            path == "open path" ~ "confounding effect",
            TRUE ~ NA_character_
        ),
        effects = factor(effects, c("true effect", "confounding effect"))
    ) |>
    ggplot(aes(x = x, y = y, xend = xend, yend = yend, color = effects, alpha = path)) +
    geom_dag_edges(aes(edge_alpha = path, edge_colour = effects), show.legend = FALSE) +
    geom_dag_point(
        data = function(.x) dplyr::filter(.x, is.na(path)),
        key_glyph = glyph
    ) +
    geom_dag_point(
        data = function(.x) dplyr::filter(.x, !is.na(path)),
        key_glyph = glyph
    ) +
    facet_wrap(vars(fct_inorder(factor(set))), nrow = 2) +
    expand_plot(
        expand_x = expansion(c(0.25, 0.25)),
        expand_y = expansion(c(0.1, 0.1))
    ) +
    theme_dag() +
    theme(
        legend.position = "top",
        legend.spacing.x = unit(8, "mm"),
        legend.text = element_text(size = rel(2.5)),
        legend.box.margin = margin(b = 20),
        strip.text = element_blank()
    ) +
    coord_cartesian(clip = "off") +
    scale_alpha_manual(
        drop = FALSE,
        values = c("open path" = 1),
        na.value = 0.5,
        breaks = "open path"
    ) +
    ggraph::scale_edge_alpha_manual(
        drop = FALSE,
        values = c("open path" = 1),
        na.value = 0.5,
        breaks = "open path"
    ) +
    scale_color_okabe_ito(
        name = NULL,
        na.value = "grey90",
        order = c(3, 6),
        breaks = c("true effect", "confounding effect")
    ) +
    scale_edge_color_okabe_ito(
        name = NULL,
        na.value = "grey90",
        order = c(3, 6),
        breaks = c("true effect", "confounding effect")
    ) +
    guides(alpha = "none", edge_alpha = "none")
```

##

```{r}
#| echo: false
#| fig-width: 9

mosquito_dag |>
  ggdag_adjustment_set(text = FALSE, use_labels = "label", node_size = 10, shadow = TRUE) +
  theme_dag() +
  facet_wrap(
    vars(set),
    labeller = as_labeller(scales::label_wrap(40))
  )
```

## Multivariable regression: what's the association?

```{r}
#| code-line-numbers: "|1-4"
#| output-location: fragment

lm(
  malaria_risk ~ net + income + health + temperature,
  data = net_data
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")
```

## {.center background-color="#23373B"}

### **model your assumptions**

## {.center background-color="#23373B"}

### counterfactual: what if <u>everyone</u> used a net vs. what if <u>no one</u> used a net

## Fit propensity score model

```{r}
#| code-line-numbers: "|2"

propensity_model <- glm(
  net ~ income + health + temperature,
  family = binomial(),
  data = net_data
)
```

## Calculate inverse probability weights

```{r}
#| code-line-numbers: "|4,6"

library(propensity)
net_data_wts <- propensity_model |>
  # predict whether a household used a net
  augment(data = net_data, type.predict = "response") |>
  # calculate inverse probability
  mutate(wts = wt_ate(.fitted, net))
```

## {.center background-color="#23373B"}

### **diagnose your model assumptions**

## What's the distribution of weights?

```{r}
#| echo: false

ggplot(net_data_wts, aes(wts)) +
  geom_histogram(color = "white", fill = "#E69F00", bins = 50) +
  theme_minimal(base_size = 20) +
  xlab("Weights")
```

## What are the weights doing?

```{r}
#| echo: false

ggplot(net_data_wts, aes(.fitted)) +
  geom_mirror_histogram(
    aes(fill = net),
    bins = 50
  ) +
  scale_y_continuous(labels = abs) +
  labs(x = "propensity score") +
  theme_minimal(base_size = 20)
```

## What are the weights doing?

```{r}
#| echo: false

ggplot(net_data_wts, aes(.fitted)) +
  geom_mirror_histogram(
    aes(group = net),
    bins = 50
  ) +
  geom_mirror_histogram(
    aes(fill = net, weight = wts),
    bins = 50,
    alpha = 0.5
  ) +
  scale_y_continuous(labels = abs) +
  labs(x = "propensity score") +
  theme_minimal(base_size = 20)
```

##

```{r}
#| echo: false

plot_df <- tidy_smd(
  net_data_wts,
  c(income, health, temperature),
  .group = net,
  .wts = wts
)
```

```{r}
#| echo: false

ggplot(
  data = plot_df |> filter(method == "observed"),
  aes(x = abs(smd), y = variable, group = method, color = method)
) +
  geom_love() +
  xlim(0, 0.4) +
  theme_minimal(base_size = 20) +
  theme(axis.title.y = element_blank())
```

##

```{r}
#| echo: false

ggplot(
  data = plot_df,
  aes(x = abs(smd), y = variable, group = method, color = method)
) +
  geom_love() +
  scale_color_manual(values = c("grey85", "#00BFC4")) +
  xlim(0, 0.4) +
  theme_minimal(base_size = 20) +
  theme(axis.title.y = element_blank())
```

## {.center background-color="#23373B"}

### **estimate the causal effects**

## Estimate causal effect with IPW

```{r}
#| code-line-numbers: "|1-2,4"

ipw_model <- lm(
  malaria_risk ~ net,
  data = net_data_wts,
  weights = wts
)

ipw_estimate <- ipw_model |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")
```

## Estimate causal effect with IPW

```{r}
ipw_estimate
```

## Let's fix our confidence intervals with robust SEs!

. . .

```{r}
#| code-line-numbers: "|3-4,6"

# also see robustbase, survey, gee, and others
library(estimatr)
ipw_model_robust <- lm_robust(
  malaria_risk ~ net,
  data = net_data_wts,
  weights = wts
)

ipw_estimate_robust <- ipw_model_robust |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")
```

## Let's fix our confidence intervals with robust SEs!

```{r}
as_tibble(ipw_estimate_robust)
```

## Let's fix our confidence intervals with the bootstrap!

. . .

```{r}
# fit ipw model for a single bootstrap sample
fit_ipw_not_quite_rightly <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # fit ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```

## {.small}

```{r}
fit_ipw <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # fit propensity score model
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .df
  )

  # calculate inverse probability weights
  .df <- propensity_model |>
    augment(type.predict = "response", data = .df) |>
    mutate(wts = wt_ate(.fitted, net))

  # fit correctly bootstrapped ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```

## Using {rsample} to bootstrap our causal effect

. . .

```{r}
#| label: boot_cache
#| code-line-numbers: "|2-3"

# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(net_data, times = 1000, apparent = TRUE) |>
  mutate(results = map(splits, fit_ipw))
```

## Using {rsample} to bootstrap our causal effect

```{r}
#| code-line-numbers: "|2"
#| output-location: slide

# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) |>
  filter(term == "netTRUE")

boot_estimate
```

## {.center}

```{r}
#| echo: false

bind_rows(
  ipw_estimate |>
    select(estimate, conf.low, conf.high) |>
    mutate(type = "ols"),
  ipw_estimate_robust |>
    select(estimate, conf.low, conf.high) |>
    mutate(type = "robust"),
  boot_estimate |>
    select(estimate = .estimate, conf.low = .lower, conf.high = .upper) |>
    mutate(type = "bootstrap")
) |>
  #  calculate CI width to sort by it
  mutate(width = conf.high - conf.low) |>
  arrange(width) |>
  #  fix the order of the model types for the plot
  mutate(type = fct_inorder(type)) |>
  ggplot(aes(x = type, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(color = "#0172B1", size = 1, fatten = 3) +
    coord_flip() +
    theme_minimal(base_size = 20) +
    theme(axis.title.y = element_blank())
```

## {.center background-color="#23373B"}

### **conduct a sensitivity analysis**

## {.center background-color="#23373B"}

### **Spoiler alert: the answer we just calculated is wrong**

## How much confounding would tip our estimate?

```{r}
#| code-line-numbers: "|2|3"
#| output-location: slide

library(tipr)
tipping_points <- tip_coef(
  boot_estimate$.upper,
  exposure_confounder_effect = 1:5
)

tipping_points |>
  select(exposure_confounder_effect, confounder_outcome_effect)
```

##

```{r}
#| echo: false
#| fig-asp: 0.9

tipping_points |>
  ggplot(aes(confounder_outcome_effect, exposure_confounder_effect)) +
  geom_line(color = "#009E73", linewidth = 1.1) +
  geom_point(fill = "#009E73", color = "white", size = 2.5, shape = 21) +
  theme_minimal(base_size = 20) +
  labs(
    x = "Confounder-Outcome Effect",
    y = "Scaled mean difference in\nconfounder between\nexposure groups"
  )
```

## What would it take to explain away our estimate?

::: nonincremental
- A confounder with a scaled mean difference of 1 between net users and
  non-users would need an effect of -11.6 on malaria risk
- If the scaled mean difference were 5, an effect of -2.3 would be enough
:::

## An unmeasured confounder: genetic resistance to malaria

::: nonincremental
1. People with this genetic resistance have a malaria risk lower by about 10
2. About 26% of people who use nets have this genetic resistance
3. About 5% of people who don't use nets have this genetic resistance
:::

## Adjust the estimate for the unmeasured confounder

```{r}
#| code-line-numbers: "|4-8"
#| output-location: slide

adjusted_estimates <- boot_estimate |>
  select(.estimate, .lower, .upper) |>
  unlist() |>
  adjust_coef_with_binary(
    exposed_confounder_prev = 0.26,
    unexposed_confounder_prev = 0.05,
    confounder_outcome_effect = -10
  )

adjusted_estimates |>
  select(effect_observed, effect_adjusted)
```

## The DAG that generated the data

```{r}
#| echo: false
#| fig-width: 7.5

mosquito_dag_full <- dagify(
  malaria_risk ~ net + income + health + temperature +
    insecticide_resistance + genetic_resistance,
  net ~ income + health + temperature + eligible +
    household + genetic_resistance,
  eligible ~ income + household,
  health ~ income,
  exposure = "net",
  outcome = "malaria_risk",
  coords = list(
    x = c(
      malaria_risk = 7,
      net = 3,
      income = 4,
      health = 5,
      temperature = 6,
      insecticide_resistance = 8.5,
      eligible = 2,
      household = 1,
      genetic_resistance = 8.5
    ),
    y = c(
      malaria_risk = 2,
      net = 2,
      income = 3,
      health = 1,
      temperature = 3,
      insecticide_resistance = 2,
      eligible = 3,
      household = 2,
      genetic_resistance = 1
    )
  ),
  labels = c(
    malaria_risk = "Risk of malaria",
    net = "Mosquito net",
    income = "Income",
    health = "Health",
    temperature = "Nighttime temperatures",
    insecticide_resistance = "Insecticide resistance",
    eligible = "Eligible for program",
    household = "Number in the household",
    genetic_resistance = "Malaria resistance"
  )
) |>
  tidy_dagitty()

plot_mosquito_dag(mosquito_dag_full)
```

## {.small}

```{r}
fit_ipw_full <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # add genetic resistance to the propensity score model
  propensity_model <- glm(
    net ~ income + health + temperature + genetic_resistance,
    family = binomial(),
    data = .df
  )

  # calculate inverse probability weights
  .df <- propensity_model |>
    augment(type.predict = "response", data = .df) |>
    mutate(wts = wt_ate(.fitted, net))

  # fit correctly bootstrapped ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```

## Re-estimate the effect with the confounder measured

```{r}
#| label: boot_cache_full
#| code-line-numbers: "|1-2|4-5"
#| output-location: slide

ipw_results_full <- bootstraps(net_data_full, times = 1000, apparent = TRUE) |>
  mutate(results = map(splits, fit_ipw_full))

boot_estimate_full <- int_t(ipw_results_full, results) |>
  filter(term == "netTRUE")

boot_estimate_full
```

## {.center background-color="#23373B"}

### **These data are simulated. The true effect of net use on malaria risk is about -10**

## {.center background-color="#23373B"}

### *Our causal effect estimate: **-10.3 (95% CI -11.1, -9.3)***

## {.center background-color="#23373B"}

### **Review the Quarto file... later!**

## Resources {background-color="#23373B"}

### [Causal Inference in R](https://www.r-causal.org/): Our book, where this example comes from. Free online.

### [Causal Inference](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/): Comprehensive text on causal inference. Free online.

### [Bootstrap confidence intervals with {rsample}](https://rsample.tidymodels.org/articles/Applications/Intervals.html)

### [R-causal](https://github.com/r-causal): Our GitHub org with R packages and examples
