---
title: "When Standard Methods Succeed"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
format: kakashi-revealjs
---

## {.center .large background-color="#23373B"}

### when correlation *is* causation

```{r}
#| label: setup
#| include: false
#| message: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(ggdag)
library(propensity)

plot_estimates <- function(data) {
  unadjusted <- lm(y ~ treatment, data = data) |>
    tidy(conf.int = TRUE) |>
    filter(term == "treatment") |>
    mutate(model = "unadjusted")

  adjusted <- lm(y ~ treatment + weight + age, data = data) |>
    tidy(conf.int = TRUE) |>
    filter(term == "treatment") |>
    mutate(model = "direct\nadjustment")

  propensity_model <- glm(
    treatment ~ weight + age,
    family = binomial(),
    data = data
  )
  data_wts <- propensity_model |>
    augment(data = data, type.predict = "response") |>
    mutate(wts = wt_ate(.fitted, treatment, exposure_type = "binary"))
  outcome_model <- lm(y ~ treatment, data = data_wts, weights = wts)
  weighted <- ipw(propensity_model, outcome_model) |>
    tidy(conf.int = TRUE) |>
    mutate(model = "inverse\nprobability\nweighting")

  bind_rows(unadjusted, adjusted, weighted) |>
    select(model, estimate, std.error, starts_with("conf")) |>
    mutate(model = fct_inorder(model)) |>
    pivot_longer(c(estimate, std.error), names_to = "statistic") |>
    mutate(
      statistic = recode_values(
        statistic,
        "estimate" ~ "estimate (95% CI)",
        "std.error" ~ "standard error"
      )
    ) |>
    ggplot(aes(value, fct_rev(model))) +
    geom_point(size = 3, color = "#0172B1") +
    geom_errorbar(
      data = \(d) filter(d, statistic == "estimate (95% CI)"),
      aes(xmin = conf.low, xmax = conf.high),
      width = 0,
      linewidth = 1,
      color = "#0172B1"
    ) +
    facet_wrap(~statistic, scales = "free_x") +
    theme_minimal(base_size = 20) +
    theme(axis.title = element_blank())
}
```

## {.center background-color="#23373B"}

### When you have no confounders and there is a linear relationship between the exposure and the outcome, that *correlation is a causal relationship*

### 😮

## {.center background-color="#23373B"}

### When you have no confounders and there is a linear relationship between the *exposure* and the outcome, that correlation is a causal relationship

### 😮

## {.center background-color="#23373B"}

### When you have no confounders and there is a linear relationship between the exposure and the *outcome*, that correlation is a causal relationship

### 😮

## {.center background-color="#23373B"}

### When you have no *confounders* and there is a linear relationship between the exposure and the outcome, that correlation is a causal relationship

### 😮

## {.center background-color="#23373B"}

### randomized controlled trials

### A/B testing

## Even in these cases, using the methods you will learn here can help! {.small background-color="#23373B"}

1. Adjusting for baseline covariates can make an estimate *more efficient*
1. Propensity score weighting *always improves precision* over no adjustment and is *equivalent in precision to direct adjustment*
1. Sometimes we are *more comfortable with the functional form of the propensity score* (predicting exposure) than the outcome model

::: {.small}
Williamson et al., *Statistics in Medicine*, 2014
:::

## Example

- **simulated** data (100 observations)  
- Treatment is **randomly** assigned 
- There are **two baseline covariates**: `age` and `weight`

## Example

```{r}
#| echo: false

set.seed(10)
n <- 100
d <- tibble(
  age = rnorm(n, 55, 20),
  weight = rnorm(n),
  treatment = rbinom(n, 1, 0.5),
  y = treatment + 0.2 * age + 0.2 * weight + rnorm(n)
)

node_times <- data.frame(
  name = c("y", "trt", "age", "wt"),
  time = c(3, 2, 1, 1)
)
dagify(
  y ~ trt + age + wt,
  coords = time_ordered_coords(node_times)
) |>
  ggdag() +
  theme_dag()
```

- **True average treatment effect**: 1

## Three ways to estimate the effect {.small}

```{r}
#| eval: false

unadjusted <- lm(y ~ treatment, data = d)

adjusted <- lm(y ~ treatment + weight + age, data = d)

propensity_model <- glm(
  treatment ~ weight + age,
  family = binomial(),
  data = d
)
d_wts <- propensity_model |>
  augment(data = d, type.predict = "response") |>
  mutate(wts = wt_ate(.fitted, treatment, exposure_type = "binary"))
outcome_model <- lm(y ~ treatment, data = d_wts, weights = wts)
weighted <- ipw(propensity_model, outcome_model)
```

::: {.small}
`ipw()` uses a sandwich estimator that also accounts for the uncertainty in the propensity score model
:::

## {.center}

```{r}
#| echo: false

plot_estimates(d)
```

## Conditional vs. marginal effects

- Direct adjustment gives a **conditional** effect: *holding the other variables constant*
- Weighting gives a **marginal** effect: *averaged over the population*
- With a continuous outcome, a linear effect, and no interactions, the two are identical

## Example

- **simulated** data (10,000 observations)  
- Treatment is **randomly** assigned 
- There are **two baseline covariates**: `age` and `weight`

## {.center}

```{r}
#| echo: false

set.seed(10)
n <- 10000
d <- tibble(
  age = rnorm(n, 55, 20),
  weight = rnorm(n),
  treatment = rbinom(n, 1, 0.5),
  y = treatment + 0.2 * age + 0.2 * weight + rnorm(n)
)
plot_estimates(d)
```

## Example

- **simulated** data (10,000 observations)
- Treatment is **not** randomly assigned 
- **Two baseline confounders**: `age` and `weight`
- The effect is **homogeneous**, so marginal and conditional effects agree

## Example

```{r}
#| echo: false

set.seed(10)
n <- 10000
d <- tibble(
  age = rnorm(n, 55, 20),
  weight = rnorm(n),
  treatment = rbinom(n, 1, 1 / (1 + exp(-(0.01 * age + 0.1 * weight)))),
  y = treatment + 0.2 * age + 0.2 * weight + rnorm(n)
)

node_times <- data.frame(
  name = c("y", "trt", "age", "wt"),
  time = c(3, 2, 1, 1)
)
dagify(
  y ~ trt + age + wt,
  trt ~ age + wt,
  coords = time_ordered_coords(node_times)
) |>
  ggdag() +
  theme_dag()
```

- **True average treatment effect**: 1

## {.center}

```{r}
#| echo: false

plot_estimates(d)
```

## Example

- **simulated** data (10,000 observations)
- The exposure is **daily updates**, the outcome is **satisfaction**
- There is **one baseline confounder**: `num_users`
- **True average treatment effect**: 0

## Example

```{r}
#| echo: false

set.seed(11)
n <- 10000
d <- tibble(
  num_users = runif(n, 1, 500),
  update_frequency = rbinom(n, 1, plogis(num_users / 100)),
  satisfaction = 70 - 0.001 * (num_users - 300)^2 - 0.001 * (num_users - 300)^3
) |>
  mutate(
    satisfaction = as.numeric(scale(satisfaction)),
    update_frequency = factor(
      update_frequency,
      labels = c("weekly", "daily")
    )
  )

ggplot(d, aes(x = num_users, y = satisfaction)) +
  geom_line(linewidth = 1, color = "#0172B1") +
  theme_minimal(base_size = 20)
```

- The relationship is **non-linear**

## The wrong functional form

```{r}
lm(satisfaction ~ update_frequency + num_users, data = d) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

- The estimate is far from the truth, which is not even contained in the confidence interval

## A flexible functional form

```{r}
lm(
  satisfaction ~ update_frequency + splines::ns(num_users, 3),
  data = d
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

- We often do not know the data-generating mechanism, but we can still fit **flexible parametric models**

## {.center background-color="#23373B"}

### the model needs the right *variables* and the right *functional form*

We need the correct functional form for both the exposure and the confounders

## {.center background-color="#23373B"}

### *time-varying* confounding

Linear regression cannot estimate these types of effects without bias, so we'll need methods like inverse probability weighting or g-computation to calculate it correctly
