---
title: "Causal Assumptions"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60,
  ggplot2.discrete.colour = ggokabeito::palette_okabe_ito(),
  ggplot2.discrete.fill = ggokabeito::palette_okabe_ito()
)
library(tidyverse)
library(broom)
library(gt)
library(ggokabeito)
library(glue)
library(halfmoon)
library(hms)
library(patchwork)
library(touringplans)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- In *Split Decision: Life Stories*, Ice-T recounts his journey from a life of
  crime to fame and success
- He contrasts it with the fate of his former crime partner, Spike, now serving
  35 years to life
- The book describes how "their paths might have very well been reversed if they
  had made different choices"

## Are they good proxies for each other?

  | Person | Observed                                 | Missing counterfactual                   |
  | ------ | ---------------------------------------- | ---------------------------------------- |
  | Ice-T  | Abandons criminal life: fame and fortune | Does one more heist: 35 years in prison  |
  | Spike  | Does one more heist: 35 years in prison  | Abandons criminal life: fame and fortune |

## They differ in more than their choices

- Ice-T's decision was not the only factor in his success: he had enough musical
  talent to make a career of it
- Spike might differ in hard-to-measure ways, which makes him a poorer proxy
  than he first seems

. . .

We must rely on statistical techniques to help construct these unobservable
counterfactuals from observed data

## Potential outcomes

- Factual and counterfactual outcomes are two realizations of potential outcomes
- Say the exposure has two levels, `x = continue` doing jewel heists or
  `x = quit`
- There are two potential outcomes, `y(continue)` and `y(quit)`, and only one
  will be realized

## The missing potential outcome

- One potential outcome is observable and one is missing; early causal inference
  methods were framed as missing data problems
- We're missing `y(continue)` for Ice-T and `y(quit)` for Spike, so we can't
  calculate their individual causal effects

## {.center background-color="#23373B"}

### **Causal assumptions allow us to imbue observed data with the meaning of potential outcomes and their analysis with causal effects**

## Does chocolate make you happier?

Suppose some happiness index exists that ranges from 1 to 10

```{r}
#| code-line-numbers: "|3-4|6"
#| output-location: fragment

data <- tibble(
  id = 1:10,
  y_chocolate = c(4, 4, 6, 5, 6, 5, 6, 7, 5, 6),
  y_vanilla = c(1, 3, 4, 5, 5, 6, 8, 6, 3, 5)
) |>
  mutate(causal_effect = y_chocolate - y_vanilla)

data |>
  summarize(
    avg_chocolate = mean(y_chocolate),
    avg_vanilla = mean(y_vanilla),
    avg_causal_effect = mean(causal_effect)
  )
```

## Both potential outcomes are known

```{r}
#| echo: false

data |>
  gt() |>
  cols_label(
    id = "ID",
    y_chocolate = "y(chocolate)",
    y_vanilla = "y(vanilla)",
    causal_effect = "y(chocolate) - y(vanilla)"
  ) |>
  tab_spanner(
    label = "Potential outcomes",
    columns = c(y_chocolate, y_vanilla)
  ) |>
  tab_spanner(
    label = "Causal effect",
    columns = causal_effect
  ) |>
  tab_options(table.font.size = "60%")
```

We can calculate every individual causal effect

## In reality, each person eats one flavor {.small}

```{r}
#| code-line-numbers: "|1|4-8|9-12|15-17"
#| output-location: fragment

set.seed(11)
data_observed <- data |>
  mutate(
    exposure = if_else(
      rbinom(n(), 1, 0.5) == 1,
      "chocolate",
      "vanilla"
    ),
    observed_outcome = case_when(
      exposure == "chocolate" ~ y_chocolate,
      exposure == "vanilla" ~ y_vanilla
    )
  )

data_observed |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

We're a little off because of the low sample size

## Only the observed outcome remains

```{r}
#| echo: false

data_observed |>
  mutate(
    y_chocolate = if_else(exposure == "chocolate", y_chocolate, NA),
    y_vanilla = if_else(exposure == "vanilla", y_vanilla, NA),
    causal_effect = NA_real_
  ) |>
  select(-observed_outcome, -exposure) |>
  gt() |>
  cols_label(
    id = "ID",
    y_chocolate = "y(chocolate)",
    y_vanilla = "y(vanilla)",
    causal_effect = "y(chocolate) - y(vanilla)"
  ) |>
  sub_missing(
    columns = c(y_chocolate, y_vanilla, causal_effect),
    missing_text = ""
  ) |>
  tab_spanner(
    label = "Potential outcomes",
    columns = c(y_chocolate, y_vanilla)
  ) |>
  tab_spanner(
    label = "Causal effect",
    columns = causal_effect
  ) |>
  tab_options(table.font.size = "60%")
```

We can't calculate the individual causal effect

## Apples-to-apples

- Most causal assumptions exist to make an apples-to-apples comparison: we want
  to compare similar individuals who serve as reasonable proxies for each
  other's counterfactuals
- Other people say cheese and chalk (UK English), apples and pears (German), or
  grandmothers and toads (Serbian)

## The three assumptions

- Unconfoundedness methods assume **exchangeability**, **positivity**, and
  **consistency**
- Knowing a method's assumptions is essential for using it correctly; these are
  sometimes called identifiability conditions

## Exchangeability

Exposure status is independent of the potential outcomes; that is, each exposure
group has the same potential outcomes on average

```{r}
#| include: false

prepare_plot_data <- function(
  data,
  pivot_prefix = "y_",
  potential_outcome_transform = \(x) paste0("potential outcome: y(", x, ")"),
  transform_exposure = identity,
  id_assignment = TRUE
) {
  if (id_assignment && !"id" %in% names(data)) {
    data <- data |> mutate(id = row_number())
  }
  data |>
    pivot_longer(
      cols = starts_with(pivot_prefix),
      names_prefix = pivot_prefix,
      names_to = "potential_outcome",
      values_to = "happiness"
    ) |>
    mutate(
      observed = if_else(
        exposure == potential_outcome,
        "observed",
        "unobserved"
      )
    ) |>
    mutate(
      potential_outcome = potential_outcome_transform(potential_outcome),
      exposure = transform_exposure(exposure)
    ) |>
    arrange(id) |>
    mutate(y_id = dense_rank(id))
}

compute_avg_labels <- function(plot_data, group_vars) {
  plot_data |>
    group_by(across(all_of(group_vars))) |>
    summarize(happiness = mean(happiness, na.rm = TRUE), .groups = "drop") |>
    mutate(y_id = min(plot_data$y_id) - 1)
}

add_avg_layers <- function(
  avg_labels,
  observed_col = "grey50",
  unobserved_col = "grey50"
) {
  list(
    geom_point(
      data = avg_labels |> filter(observed == "unobserved"),
      mapping = aes(x = happiness, y = y_id),
      size = 4,
      shape = 23,
      fill = "white",
      color = unobserved_col,
      inherit.aes = FALSE
    ),
    geom_point(
      data = avg_labels |> filter(observed == "observed"),
      mapping = aes(x = happiness, y = y_id),
      size = 4,
      shape = 23,
      fill = observed_col,
      color = observed_col,
      inherit.aes = FALSE
    )
  )
}

po_theme <- theme(
  panel.border = element_rect(color = "grey40", fill = NA, linewidth = 0.8),
  axis.title.y = element_blank()
)
```

```{r}
#| echo: false
#| fig-width: 8

plot_data <- data_observed |>
  select(starts_with("y"), exposure) |>
  mutate(id = row_number()) |>
  prepare_plot_data(
    pivot_prefix = "y_",
    potential_outcome_transform = \(x) paste0("potential outcome: y(", x, ")"),
    transform_exposure = \(exp) {
      if_else(
        exp == "vanilla",
        "actually ate vanilla",
        "actually ate chocolate"
      )
    },
    id_assignment = FALSE
  )

avg_labels <- compute_avg_labels(
  plot_data,
  c("potential_outcome", "exposure", "observed")
) |>
  mutate(
    exposure_lbl = str_replace_all(exposure, "actually ate ", ""),
    po_lbl = str_replace_all(potential_outcome, "potential outcome: ", ""),
    label = glue("Avg {po_lbl}\n({exposure_lbl} group, {observed})") |>
      str_wrap(19)
  )

id_annotation <- plot_data |>
  filter(id == 3) |>
  mutate(
    label = glue("Potential outcomes\nfor ID 3 ({observed})") |> str_wrap(15)
  )

exchangeability_annotation <- tibble(
  x = 5.43,
  xend = 5.33,
  y = 1,
  yend = 0.5,
  potential_outcome = "potential outcome: y(chocolate)",
  label = str_wrap(
    "For exchangeability to hold, these\ngroup averages should be similar",
    19
  )
)

ggplot(plot_data, aes(happiness, y_id, color = observed, shape = observed)) +
  geom_point(aes(fill = observed), size = 3, alpha = 0.8) +
  add_avg_layers(
    avg_labels,
    observed_col = ggokabeito::palette_okabe_ito(1),
    unobserved_col = ggokabeito::palette_okabe_ito(2)
  ) +
  geom_curve(
    data = id_annotation,
    mapping = aes(
      x = happiness + 2.5,
      xend = happiness + 0.5,
      y = y_id + 2,
      yend = y_id
    ),
    curvature = -0.2,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = id_annotation,
    mapping = aes(x = happiness + 2, y = y_id + 1.5, label = label),
    hjust = 0,
    inherit.aes = FALSE,
    color = "grey40",
    size = 3.75,
    linewidth = NA
  ) +
  geom_curve(
    data = exchangeability_annotation,
    mapping = aes(x = x + 0.5, xend = xend + 0.2, y = y + 0.5, yend = yend),
    curvature = 0.1,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = exchangeability_annotation,
    mapping = aes(x = x + 0.05, y = y, label = label),
    inherit.aes = FALSE,
    hjust = "left",
    nudge_x = 0.5,
    color = "grey40",
    size = 4,
    linewidth = NA
  ) +
  facet_wrap(~potential_outcome) +
  scale_y_continuous(
    breaks = c(unique(plot_data$y_id), min(plot_data$y_id) - 1),
    labels = c(unique(plot_data$id), expression(bold("Avg")))
  ) +
  scale_shape_manual(
    name = NULL,
    values = c(19, 21)
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(
      "observed" = ggokabeito::palette_okabe_ito(1),
      "unobserved" = "white"
    )
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      "observed" = ggokabeito::palette_okabe_ito(1),
      "unobserved" = ggokabeito::palette_okabe_ito(2)
    )
  ) +
  scale_x_continuous(
    breaks = seq(0, 12, by = 2.5),
    limits = c(NA, 12)
  ) +
  coord_cartesian(clip = "off") +
  po_theme
```

## What if we mixed up the labels? {.small}

Say we accidentally gave chocolate to the "vanilla" group and vanilla to the
"chocolate" group

```{r}
#| code-line-numbers: "|1-3|7|14-16"
#| output-location: fragment

mix_up <- function(flavor) {
  if_else(flavor == "chocolate", "vanilla", "chocolate")
}

data_observed <- data_observed |>
  mutate(
    exposure = mix_up(exposure),
    observed_outcome = case_when(
      exposure == "chocolate" ~ y_chocolate,
      exposure == "vanilla" ~ y_vanilla
    )
  )

data_observed |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

Flavor assignment is independent of the potential outcomes

## What if participants chose their own flavor? {.small}

```{r}
#| code-line-numbers: "|4|5-17|24-26"
#| output-location: fragment

set.seed(113)
data_observed_exch <- data |>
  mutate(
    prefer_chocolate = y_chocolate > y_vanilla,
    exposure = case_when(
      # people choose the flavor they prefer 80% of the time
      prefer_chocolate ~ if_else(
        rbinom(n(), 1, 0.8) == 1,
        "chocolate",
        "vanilla"
      ),
      !prefer_chocolate ~ if_else(
        rbinom(n(), 1, 0.8) == 1,
        "vanilla",
        "chocolate"
      )
    ),
    observed_outcome = case_when(
      exposure == "chocolate" ~ y_chocolate,
      exposure == "vanilla" ~ y_vanilla
    )
  )

data_observed_exch |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

Now it seems like vanilla makes you happier!

## The groups are no longer exchangeable

```{r}
#| echo: false
#| fig-width: 8

data_observed_exch |>
  select(starts_with("y"), exposure) |>
  pivot_longer(
    starts_with("y"),
    names_prefix = "y_",
    names_to = "potential_outcome",
    values_to = "happiness"
  ) |>
  mutate(
    observed = if_else(exposure == potential_outcome, "observed", "unobserved"),
    potential_outcome = paste0("potential outcome: y(", potential_outcome, ")"),
    exposure = if_else(
      exposure == "vanilla",
      "actually ate\nvanilla",
      "actually ate\nchocolate"
    )
  ) |>
  ggplot(aes(
    happiness,
    exposure,
    color = observed,
    fill = observed,
    shape = observed
  )) +
  stat_summary(
    fun = "mean",
    size = 3.5,
    geom = "point",
    shape = 23,
    position = position_nudge(y = 0.033)
  ) +
  stat_summary(
    fun = "mean",
    geom = "text",
    aes(label = round(after_stat(x), 1)),
    vjust = 1.8,
    show.legend = FALSE
  ) +
  facet_wrap(~potential_outcome) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "grey40", fill = NA, linewidth = 0.8),
    axis.title.y = element_blank()
  ) +
  labs(
    y = "actual exposure",
    color = NULL,
    shape = NULL,
    fill = NULL
  ) +
  coord_cartesian(clip = "off") +
  scale_shape_manual(values = c(19, 21)) +
  scale_fill_manual(
    values = c(
      observed = ggokabeito::palette_okabe_ito(1),
      unobserved = "white"
    )
  ) +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5), limits = c(-2, 12))
```

Preference influences both flavor and happiness

## Conditional exchangeability

Exchangeability can hold within preference groups

```{r}
#| echo: false
#| fig-width: 8

data_observed_exch |>
  mutate(
    prefer_chocolate = if_else(
      prefer_chocolate,
      "prefers\nchocolate",
      "prefers\nvanilla"
    )
  ) |>
  pivot_longer(
    starts_with("y"),
    names_prefix = "y_",
    names_to = "potential_outcome",
    values_to = "happiness"
  ) |>
  mutate(
    observed = if_else(exposure == potential_outcome, "observed", "unobserved"),
    potential_outcome = paste0("potential outcome: y(", potential_outcome, ")"),
    exposure = if_else(
      exposure == "vanilla",
      "actually ate\nvanilla",
      "actually ate\nchocolate"
    )
  ) |>
  ggplot(aes(
    happiness,
    exposure,
    color = observed,
    fill = observed,
    shape = observed
  )) +
  stat_summary(
    fun = "mean",
    size = 3.5,
    geom = "point",
    shape = 23,
    position = position_nudge(y = 0.033)
  ) +
  stat_summary(
    fun = "mean",
    geom = "text",
    aes(label = round(after_stat(x), 1)),
    vjust = 1.8,
    show.legend = FALSE
  ) +
  facet_grid(prefer_chocolate ~ potential_outcome) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "grey40", fill = NA, linewidth = 0.8),
    axis.title.y = element_blank()
  ) +
  labs(
    y = "actual exposure",
    color = NULL,
    shape = NULL,
    fill = NULL
  ) +
  coord_cartesian(clip = "off") +
  scale_shape_manual(values = c(19, 21)) +
  scale_fill_manual(
    values = c(
      observed = ggokabeito::palette_okabe_ito(1),
      unobserved = "white"
    )
  ) +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5), limits = c(-2, 12))
```

. . .

But few values fall in each combination

## Positivity

- Everyone has a non-zero probability of each exposure level
- **Stochastic violations** are chance occurrences with no observations for a
  given exposure level
- **Structural violations** happen when an individual cannot receive an exposure
  level by definition

## A stochastic violation

Participants choose the flavor they prefer 80% of the time, as before

```{r}
#| include: false

set.seed(1)
data_observed_pos <- data |>
  mutate(
    prefer_chocolate = y_chocolate > y_vanilla,
    exposure = case_when(
      prefer_chocolate ~ if_else(
        rbinom(n(), 1, 0.8) == 1,
        "chocolate",
        "vanilla"
      ),
      !prefer_chocolate ~ if_else(
        rbinom(n(), 1, 0.8) == 1,
        "vanilla",
        "chocolate"
      )
    ),
    observed_outcome = case_when(
      exposure == "chocolate" ~ y_chocolate,
      exposure == "vanilla" ~ y_vanilla
    )
  )
```

```{r}
#| output-location: fragment

data_observed_pos |>
  count(prefer_chocolate, exposure) |>
  complete(
    prefer_chocolate,
    exposure = c("chocolate", "vanilla"),
    fill = list(n = 0)
  )
```

. . .

Positivity must hold within the covariates needed for exchangeability

## A structural violation: an allergy to vanilla {.small}

Everyone is randomized to a flavor, as before

```{r}
#| include: false

set.seed(11)
data_observed_struc <- data |>
  mutate(
    exposure = if_else(
      rbinom(n(), 1, 0.5) == 1,
      "chocolate",
      "vanilla"
    )
  )
```

```{r}
#| code-line-numbers: "|4|6-7|15-17"
#| output-location: fragment

set.seed(1)
data_observed_struc <- data_observed_struc |>
  mutate(
    allergy = rbinom(n(), 1, 0.3) == 1,
    # `y_vanilla` is impossible for those with an allergy
    exposure = if_else(allergy, "chocolate", exposure),
    y_vanilla = if_else(allergy, NA, y_vanilla),
    observed_outcome = case_when(
      allergy ~ y_chocolate,
      exposure == "chocolate" ~ y_chocolate,
      exposure == "vanilla" ~ y_vanilla
    )
  )

data_observed_struc |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

For those with a vanilla allergy, `y(vanilla)` is not defined

## Undefined potential outcomes

```{r}
#| echo: false
#| warning: false
#| fig-width: 8

plot_data <- data_observed_struc |>
  mutate(is_missing_y_vanilla = is.na(y_vanilla)) |>
  select(id, starts_with("y"), exposure, is_missing_y_vanilla) |>
  prepare_plot_data(
    pivot_prefix = "y_",
    potential_outcome_transform = \(x) paste0("y(", x, ")"),
    transform_exposure = \(exp) {
      if_else(
        exp == "vanilla",
        "actually ate vanilla",
        "actually ate chocolate"
      )
    },
    id_assignment = FALSE
  )

avg_labels <- compute_avg_labels(
  plot_data,
  c("potential_outcome", "exposure", "observed")
) |>
  mutate(
    exposure_lbl = str_replace_all(exposure, "actually ate ", ""),
    po_lbl = str_replace_all(potential_outcome, "potential outcome: ", ""),
    label = glue("Avg {po_lbl}\n({exposure_lbl} group, {observed})") |>
      str_wrap(19)
  )

missing_points <- plot_data |>
  filter(is_missing_y_vanilla, exposure == "actually ate chocolate") |>
  select(happiness, y_id) |>
  drop_na()

missing_annotation <- tibble(
  x = max(missing_points$happiness, na.rm = TRUE) + 1,
  y = max(missing_points$y_id, na.rm = TRUE) - 1,
  label = "Missing y(vanilla)\ncounterpart",
  potential_outcome = "y(chocolate)"
)

missing_arrows <- missing_points |>
  mutate(
    xend = happiness + 0.3,
    yend = y_id - c(-0.3, 0, 0.3),
    x = missing_annotation$x,
    y = missing_annotation$y,
    potential_outcome = "y(chocolate)"
  )

ggplot(
  plot_data,
  aes(
    x = happiness,
    y = y_id,
    color = is_missing_y_vanilla,
    fill = is_missing_y_vanilla,
    shape = observed
  )
) +
  geom_point(
    data = plot_data |> filter(observed == "observed"),
    mapping = aes(fill = is_missing_y_vanilla),
    size = 3,
    shape = 21,
    alpha = 0.8
  ) +
  geom_point(
    data = plot_data |> filter(observed == "unobserved"),
    mapping = aes(x = happiness, y = y_id),
    size = 3,
    shape = 21,
    fill = "white",
    color = "grey70",
    alpha = 0.8,
    inherit.aes = FALSE
  ) +
  add_avg_layers(avg_labels) +
  geom_curve(
    data = missing_arrows,
    mapping = aes(x = x, xend = xend, y = y, yend = yend),
    curvature = 0,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = missing_annotation,
    mapping = aes(x = x, y = y, label = label),
    hjust = 0,
    inherit.aes = FALSE,
    color = "grey40",
    size = 4.5,
    linewidth = NA
  ) +
  facet_wrap(~potential_outcome) +
  scale_y_continuous(
    breaks = c(unique(plot_data$y_id), min(plot_data$y_id) - 1),
    labels = c(unique(plot_data$id), expression(bold("Avg")))
  ) +
  scale_shape_manual(values = c(19, 21)) +
  scale_fill_manual(
    name = NULL,
    values = c("TRUE" = ggokabeito::palette_okabe_ito(7)),
    labels = c("TRUE" = "Missing y(vanilla) counterpart"),
    na.value = "grey80"
  ) +
  scale_color_manual(
    name = NULL,
    values = c("TRUE" = ggokabeito::palette_okabe_ito(7)),
    labels = c("TRUE" = "Missing y(vanilla) counterpart"),
    na.value = "grey80"
  ) +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5), limits = c(NA, 12)) +
  coord_cartesian(clip = "off") +
  po_theme
```

. . .

We can collect more data, use a statistical model to extrapolate, add
eligibility criteria, or change the estimand

## Consistency

- The causal question you claim you are answering is consistent with the one you
  are actually answering
- The potential outcome of a given treatment value equals the value we actually
  observe when someone is assigned that treatment value

. . .

It seems almost silly when you say it. What else would it be?

## Two ways consistency fails

- **Poorly-defined exposure**: multiple treatment versions exist, from surgeries
  to income to education
- **Interference**: the outcome for any subject depends on another subject's
  exposure

## Two containers of chocolate, one spoiled {.small}

Everyone has a third potential outcome, `y_spoiled_chocolate`, which is 0

```{r}
#| include: false

data <- tibble(
  id = 1:10,
  y_spoiled_chocolate = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  y_chocolate = c(4, 4, 6, 5, 6, 5, 6, 7, 5, 6),
  y_vanilla = c(1, 3, 4, 5, 5, 6, 8, 6, 3, 5)
) |>
  mutate(causal_effect = y_chocolate - y_vanilla)
```

```{r}
#| code-line-numbers: "|4-8|14-15|18-20"
#| output-location: fragment

set.seed(11)
data_observed_poorly_defined <- data |>
  mutate(
    exposure_unobserved = case_when(
      rbinom(n(), 1, 0.25) == 1 ~ "chocolate (spoiled)",
      rbinom(n(), 1, 0.25) == 1 ~ "chocolate",
      .default = "vanilla"
    ),
    observed_outcome = case_when(
      exposure_unobserved == "chocolate (spoiled)" ~ y_spoiled_chocolate,
      exposure_unobserved == "chocolate" ~ y_chocolate,
      exposure_unobserved == "vanilla" ~ y_vanilla
    ),
    # both containers are recorded as "chocolate"
    exposure = if_else(exposure_unobserved == "vanilla", "vanilla", "chocolate")
  )

data_observed_poorly_defined |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

The true causal effect of unspoiled chocolate is 0.8, but our estimate is
about -1.9

## The chocolate group is a mixture

```{r}
#| echo: false
#| fig-width: 8

plot_data <- data_observed_poorly_defined |>
  mutate(is_spoiled = exposure_unobserved == "chocolate (spoiled)") |>
  pivot_longer(
    cols = starts_with("y"),
    names_prefix = "y_",
    names_to = "potential_outcome",
    values_to = "happiness"
  ) |>
  filter(
    !(is_spoiled & potential_outcome == "chocolate"),
    !(!is_spoiled & potential_outcome == "spoiled_chocolate")
  ) |>
  mutate(
    potential_outcome = case_when(
      potential_outcome == "spoiled_chocolate" ~ "chocolate",
      TRUE ~ potential_outcome
    ),
    observed = if_else(exposure == potential_outcome, "observed", "unobserved"),
    potential_outcome = paste0("y(", potential_outcome, ")")
  ) |>
  mutate(observed = factor(observed, levels = c("observed", "unobserved"))) |>
  arrange(id) |>
  mutate(y_id = dense_rank(id))

avg_labels <- compute_avg_labels(
  plot_data,
  c("potential_outcome", "exposure", "observed")
)

spoiled_annotation <- plot_data |>
  filter(is_spoiled, potential_outcome == "y(chocolate)") |>
  slice(1) |>
  mutate(label = "Their chocolate was spoiled")

spoiled_arrows <- plot_data |>
  filter(is_spoiled, potential_outcome == "y(chocolate)") |>
  mutate(
    xend = happiness + c(0.35, 0.05),
    yend = y_id - c(0, 0.2),
    x = spoiled_annotation$happiness + 1.9,
    y = spoiled_annotation$y_id + 0.5
  )

flawed_avg_annotation <- avg_labels |>
  filter(
    potential_outcome == "y(chocolate)",
    exposure == "chocolate",
    observed == "observed"
  ) |>
  mutate(
    label = "This is an average of\nboth potential outcomes",
    x = happiness + 2.25,
    y = y_id + 0.9,
    happiness = happiness + 0.25,
    y_id = y_id + 0.15
  )

unspoiled_annotation <- plot_data |>
  filter(
    !is_spoiled,
    exposure == "chocolate",
    potential_outcome == "y(chocolate)",
    observed == "observed"
  ) |>
  slice(1) |>
  mutate(
    label = "But theirs\nwasn't",
    x = happiness + 2,
    y = y_id
  )

unspoiled_arrows <- plot_data |>
  filter(
    !is_spoiled,
    exposure == "chocolate",
    potential_outcome == "y(chocolate)",
    observed == "observed"
  ) |>
  mutate(
    xend = happiness + 0.35,
    yend = y_id - c(0.1, 0.2),
    x = unspoiled_annotation$x,
    y = unspoiled_annotation$y
  )

ggplot(plot_data, aes(x = happiness, y = y_id)) +
  geom_point(
    data = plot_data |> filter(!is_spoiled, observed == "observed"),
    size = 3,
    shape = 21,
    fill = "grey50",
    color = "grey50",
    alpha = 0.8
  ) +
  geom_point(
    data = plot_data |> filter(!is_spoiled, observed == "unobserved"),
    size = 3,
    shape = 21,
    fill = "white",
    color = "grey50",
    alpha = 0.8
  ) +
  geom_point(
    data = plot_data |>
      filter(
        is_spoiled,
        potential_outcome == "y(chocolate)",
        observed == "observed"
      ),
    size = 3,
    shape = 21,
    fill = ggokabeito::palette_okabe_ito(7),
    color = ggokabeito::palette_okabe_ito(7),
    alpha = 0.8
  ) +
  geom_point(
    data = plot_data |>
      filter(
        is_spoiled,
        potential_outcome == "y(chocolate)",
        observed == "unobserved"
      ),
    size = 3,
    shape = 21,
    fill = "white",
    color = ggokabeito::palette_okabe_ito(7),
    alpha = 0.8
  ) +
  add_avg_layers(avg_labels) +
  geom_curve(
    data = spoiled_arrows,
    mapping = aes(x = x, xend = xend, y = y, yend = yend),
    curvature = -0.2,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = spoiled_annotation,
    mapping = aes(x = happiness + 2, y = y_id + 0.5, label = label),
    hjust = 0,
    inherit.aes = FALSE,
    color = "grey40",
    size = 4,
    linewidth = NA
  ) +
  geom_curve(
    data = flawed_avg_annotation,
    mapping = aes(x = x, xend = happiness, y = y, yend = y_id),
    curvature = 0.2,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = flawed_avg_annotation,
    mapping = aes(x = x, y = y, label = label),
    hjust = 0,
    inherit.aes = FALSE,
    color = "grey40",
    size = 4,
    linewidth = NA
  ) +
  geom_curve(
    data = unspoiled_arrows,
    mapping = aes(x = x, xend = xend, y = y, yend = yend),
    curvature = -0.2,
    arrow = arrow(length = unit(0.02, "npc")),
    inherit.aes = FALSE,
    color = "grey40"
  ) +
  geom_label(
    data = unspoiled_annotation,
    mapping = aes(x = x, y = y, label = label),
    hjust = 0,
    inherit.aes = FALSE,
    color = "grey40",
    size = 4,
    linewidth = NA
  ) +
  facet_wrap(~potential_outcome) +
  scale_y_continuous(
    breaks = c(unique(plot_data$y_id), min(plot_data$y_id) - 1),
    labels = c(unique(plot_data$id), expression(bold("Avg")))
  ) +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5), limits = c(NA, 12)) +
  labs(y = "actual exposure") +
  coord_cartesian(clip = "off") +
  po_theme
```

We're treating fresh and spoiled chocolate as the same exposure; the same goes
for brands, timing, and portion size

## Separate the potential outcomes

Say we tested the containers after the experiment and know which ones were
spoiled

```{r}
#| output-location: fragment

data_observed_poorly_defined |>
  group_by(exposure_unobserved) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

Now we're back to the right answer because we've correctly separated the
potential outcomes

## The data match potential outcomes

```{r}
#| echo: false
#| fig-width: 8

plot_data <- data_observed_poorly_defined |>
  pivot_longer(
    cols = starts_with("y"),
    names_prefix = "y_",
    names_to = "potential_outcome",
    values_to = "happiness"
  ) |>
  mutate(
    potential_outcome = if_else(
      potential_outcome == "spoiled_chocolate",
      "chocolate (spoiled)",
      potential_outcome
    ),
    observed = if_else(
      exposure_unobserved == potential_outcome,
      "observed",
      "unobserved"
    ),
    potential_outcome = if_else(
      potential_outcome == "chocolate (spoiled)",
      "spoiled_chocolate",
      potential_outcome
    ),
    potential_outcome = paste0("y(", potential_outcome, ")")
  ) |>
  arrange(id) |>
  mutate(y_id = dense_rank(id))

avg_labels <- compute_avg_labels(
  plot_data,
  c("potential_outcome", "exposure_unobserved", "observed")
)

ggplot(
  plot_data,
  aes(x = happiness, y = y_id, color = exposure_unobserved)
) +
  geom_point(
    data = plot_data |> filter(observed == "observed"),
    mapping = aes(fill = exposure_unobserved),
    size = 3,
    shape = 21,
    alpha = 0.8
  ) +
  geom_point(
    data = plot_data |> filter(observed == "unobserved"),
    size = 3,
    shape = 21,
    fill = "white",
    alpha = 0.8
  ) +
  geom_point(
    data = avg_labels |> filter(observed == "unobserved"),
    aes(x = happiness, y = y_id, color = exposure_unobserved),
    size = 4,
    shape = 23,
    fill = "white",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = avg_labels |> filter(observed == "observed"),
    aes(
      x = happiness,
      y = y_id,
      fill = exposure_unobserved,
      color = exposure_unobserved
    ),
    size = 4,
    shape = 23,
    inherit.aes = FALSE
  ) +
  facet_wrap(~potential_outcome) +
  scale_y_continuous(
    breaks = c(unique(plot_data$y_id), min(plot_data$y_id) - 1),
    labels = c(unique(plot_data$id), expression(bold("Avg")))
  ) +
  scale_fill_manual(values = ggokabeito::palette_okabe_ito(c(1, 2, 7))) +
  scale_color_manual(values = ggokabeito::palette_okabe_ito(c(1, 2, 7))) +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5), limits = c(NA, 12)) +
  labs(
    y = "True exposure",
    color = NULL,
    fill = NULL
  ) +
  coord_cartesian(clip = "off") +
  po_theme
```

. . .

There will almost always be some consistency violation; the question is whether
it is meaningful

## Interference

- An individual's exposure impacts another's potential outcome, as when
  someone's vaccination affects someone else's risk
- Here, each person has a partner, and having a partner with a different flavor
  increases happiness by two units

## Each person has a partner

```{r}
#| code-line-numbers: "|8-9"

data <- tibble(
  id = 1:10,
  partner_id = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5),
  y_chocolate_chocolate = c(4, 4, 6, 5, 6, 5, 6, 7, 5, 6),
  y_vanilla_vanilla = c(1, 3, 4, 5, 5, 6, 8, 6, 3, 5)
) |>
  mutate(
    y_chocolate_vanilla = y_chocolate_chocolate + 2,
    y_vanilla_chocolate = y_vanilla_vanilla + 2
  )
```

## Randomize each individual and their partner {.small}

```{r}
#| code-line-numbers: "|4-9|10-18|21-23"
#| output-location: fragment

set.seed(37)
data_observed_interf <- data |>
  mutate(
    exposure = if_else(rbinom(n(), 1, 0.5) == 1, "chocolate", "vanilla"),
    exposure_partner = if_else(
      rbinom(n(), 1, 0.5) == 1,
      "chocolate",
      "vanilla"
    ),
    observed_outcome = case_when(
      exposure == "chocolate" & exposure_partner == "chocolate" ~
        y_chocolate_chocolate,
      exposure == "chocolate" & exposure_partner == "vanilla" ~
        y_chocolate_vanilla,
      exposure == "vanilla" & exposure_partner == "chocolate" ~
        y_vanilla_chocolate,
      exposure == "vanilla" & exposure_partner == "vanilla" ~ y_vanilla_vanilla
    )
  )

data_observed_interf |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

Each person has three counterfactuals now, and these averages don't estimate any
of them

## Specify the potential outcomes

```{r}
#| output-location: fragment

data_observed_interf |>
  group_by(exposure, exposure_partner) |>
  summarise(avg_outcome = mean(observed_outcome), .groups = "drop")
```

. . .

One way to combat interference is to change the unit: randomize the
partnerships, a cluster randomized trial

## Randomize the partnerships {.small}

```{r}
#| code-line-numbers: "|2-5|10|18-20"
#| output-location: fragment

set.seed(11)
partners <- tibble(
  partner_id = 1:5,
  exposure = if_else(rbinom(5, 1, 0.5) == 1, "chocolate", "vanilla")
)

partners_observed <- data |>
  left_join(partners, by = "partner_id") |>
  mutate(
    exposure_partner = exposure,
    observed_outcome = case_when(
      exposure == "chocolate" & exposure_partner == "chocolate" ~
        y_chocolate_chocolate,
      exposure == "vanilla" & exposure_partner == "vanilla" ~ y_vanilla_vanilla
    )
  )

partners_observed |>
  group_by(exposure) |>
  summarise(avg_outcome = mean(observed_outcome))
```

. . .

There is no interference between different partner sets

## {.center background-color="#23373B"}

### **Malaria nets, revisited**

## Malaria nets and the assumptions

- **Exchangeability**: we saw what can happen with an unmeasured confounder,
  genetic resistance to malaria
- **Positivity**: a violation would occur if any household would always or never
  use a bed net, or if no one of low economic status in cold weather ever used
  one

## Consistency and interference

- **Consistency**: a "bed net" can mean many things; does the manufacturer
  matter? The type of insecticide? The amount?
- Interference is a real problem for bed-net research, though households as the
  unit reduce some of it

## {.center background-color="#23373B"}

### **When do study designs support causal inference?**

## Randomized trials

- In the limit, we expect exchangeability: randomization is the only cause of
  exposure
- Randomization guarantees positivity so long as no one has a deterministic
  exposure
- Trials help, but do not guarantee, consistency; interference can still happen

## Trials in practice

- In actual trials, exchangeability can fail by chance, people do not adhere to
  their assigned exposure, and people drop out
- Observational studies have none of a trial's guarantees, even ideal ones

. . .

A randomized design is sometimes called an A/B test

## Assumptions solved by study design

  | Assumption                          | Ideal randomized trial | Realistic randomized trial | Observational study |
  | ----------------------------------- | ---------------------- | -------------------------- | ------------------- |
  | Consistency (well-defined exposure) | 😄                     | 🤷                         | 🤷                  |
  | Consistency (no interference)       | 🤷                     | 🤷                         | 🤷                  |
  | Positivity                          | 😄                     | 😄                         | 🤷                  |
  | Exchangeability                     | 😄                     | 🤷                         | 🤷                  |

😄 solved by default, 🤷 solvable but not solved by default

## The elements of a target trial

A clear causal question maps to a protocol

- **Eligibility criteria**: who or what is included?
- **Exposure definition**: what precise exposure?
- **Assignment procedures**: how are units assigned?
- **Follow-up period**: when does it start and end?

## More elements of the protocol

- **Outcome definition**: what precise outcomes will be measured?
- **Causal contrast of interest**: which causal estimand will be estimated?
- **Analysis plan**: what procedures will estimate the causal contrast?

## How the protocol helps

- Eligibility criteria prevent positivity violations by excluding those who can
  never have one of the exposures
- A precise exposure definition improves consistency and helps control
  interference

## What randomization adds

- Randomized assignment meets exchangeability and positivity, and can prevent
  interference

## Emulating target trials

- We often cannot randomize: time, money, ethics
- You specify a target trial, the randomized trial you would run, and emulate it
  with observational data
- The target trial is usually pragmatic: no blinding

. . .

This is step 1 of the whole game: specify your causal question, for example as a
target trial

## Extra Magic Mornings

- Guests staying at a Walt Disney World resort hotel could enter the park during
  "Extra Magic Hours"
- Does an Extra Magic Morning change the average posted wait time for the Seven
  Dwarfs Mine Train between 9 AM and 10 AM?

## The Seven Dwarfs data

You will work with these data in the propensity score exercises

```{r}
#| output-location: fragment

seven_dwarfs_9 <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9)

seven_dwarfs_9 |>
  count(park_extra_magic_morning)
```

. . .

Data can't solve the problem of the unverifiable assumptions, but it can still
provide valuable information

## Positivity: time of park close

Exchangeability needs better tools, which we'll see later, and we've addressed
consistency by being specific about Extra Magic Mornings

```{r}
#| echo: false
#| fig-asp: 0.45

ggplot(
  seven_dwarfs_9,
  aes(
    x = factor(park_close),
    group = factor(park_extra_magic_morning),
    fill = factor(park_extra_magic_morning)
  )
) +
  geom_bar(position = "fill", alpha = 0.8) +
  labs(
    fill = "Extra Magic Morning",
    x = "Time of Park Close"
  ) +
  theme(panel.grid.major.x = element_blank())
```

None of the days that closed at 16:30 or 21:00 had Extra Magic Mornings; ticket
season, by contrast, shows no violations

## Which days are missing a group?

```{r}
#| output-location: fragment

seven_dwarfs_9 |>
  count(park_close, park_extra_magic_morning) |>
  complete(
    park_close,
    park_extra_magic_morning,
    fill = list(n = 0)
  ) |>
  filter(park_close %in% parse_hm(c("16:30", "21:00")))
```

. . .

Only background knowledge can tell us whether a violation is stochastic or
structural

## Historic high temperature

```{r}
#| echo: false
#| fig-asp: 0.4

ggplot(
  seven_dwarfs_9,
  aes(
    x = park_temperature_high,
    group = factor(park_extra_magic_morning),
    fill = factor(park_extra_magic_morning)
  )
) +
  geom_mirror_histogram(bins = 20, alpha = 0.8) +
  scale_y_continuous(labels = abs) +
  labs(
    fill = "Extra Magic Morning",
    x = "Historic high temperature (F)"
  )
```

```{r}
#| output-location: fragment

seven_dwarfs_9 |>
  filter(park_temperature_high < 60) |>
  count(park_extra_magic_morning)
```

. . .

Restricting to warmer days would change the question and restrict which days the
conclusions cover

## Checking several variables at once

```{r}
#| echo: false
#| fig-width: 9
#| fig-asp: 0.5

prop_exposed <- seven_dwarfs_9 |>
  mutate(
    park_temperature_high_bin = cut(park_temperature_high, breaks = 3),
    park_close_bin = case_when(
      hour(park_close) < 19 & hour(park_close) > 12 ~ "(1) early",
      hour(park_close) >= 19 & hour(park_close) < 24 ~ "(2) standard",
      hour(park_close) >= 24 | hour(park_close) < 12 ~ "(3) late"
    )
  ) |>
  group_by(
    park_close_bin,
    park_temperature_high_bin,
    park_ticket_season
  ) |>
  summarize(
    prop_exposed = mean(park_extra_magic_morning),
    .groups = "drop"
  ) |>
  complete(
    park_close_bin,
    park_temperature_high_bin,
    park_ticket_season,
    fill = list(prop_exposed = 0)
  )

prop_exposed |>
  ggplot(
    aes(
      x = park_close_bin,
      y = park_temperature_high_bin,
      fill = prop_exposed
    )
  ) +
  geom_tile() +
  scale_fill_viridis_c(begin = 0.1, end = 0.9, breaks = c(0, 0.5, 1)) +
  facet_wrap(~park_ticket_season) +
  labs(
    y = "Historic High Temperature (F)",
    x = "Magic Kingdom Park Close Time",
    fill = "Proportion of\nDays Exposed"
  ) +
  theme(panel.grid = element_blank())
```

Nine combinations were never exposed, and one was always exposed

. . .

The only cold peak days in Florida are over Christmas and New Year, when there
historically were always Extra Magic Hours

## Overlap and balance

We look for two things: overlap and balance

```{r}
#| echo: false
#| fig-width: 9
#| fig-asp: 0.45

set.seed(2025)

make_model <- function(n = 1000, intercept, slope) {
  df <- tibble(
    z = rnorm(n),
    exposure = rbinom(n, 1, plogis(intercept + slope * z))
  )
  glm(exposure ~ z, data = df, family = binomial)
}

models_overlap <- list(
  good = make_model(intercept = 0, slope = 0.5),
  moderate = make_model(intercept = 0, slope = 1.5),
  poor = make_model(intercept = 0, slope = 3.0)
)

plot_ps <- function(.x, .y, title = "Overlap") {
  .x |>
    augment(type.predict = "response") |>
    ggplot(aes(.fitted, fill = factor(exposure))) +
    geom_mirror_histogram(bins = 50) +
    scale_y_continuous(labels = abs) +
    labs(
      x = "propensity score",
      fill = "exposure",
      title = paste(str_to_title(.y), title)
    ) +
    theme(legend.position = "none")
}

plots_overlap <- imap(models_overlap, plot_ps)

models_balance <- list(
  good = make_model(intercept = 0.0, slope = 1.5),
  moderate = make_model(intercept = log(0.25 / 0.75), slope = 1.5),
  poor = make_model(intercept = log(0.10 / 0.90), slope = 1.5)
)

plots_balance <- imap(models_balance, plot_ps, title = "Balance")

plots_balance$moderate <- plots_balance$moderate +
  theme(legend.position = "bottom")

(plots_overlap$good + plots_overlap$moderate + plots_overlap$poor) /
  (plots_balance$good + plots_balance$moderate + plots_balance$poor)
```

Overlap, or common support, relates to positivity

. . .

Balance relates to exchangeability: we should be able to reassign the two groups
and still get the right answer

. . .

The x-axis is the probability of exposure given the covariates: the propensity
score, which we take up next

## *Your Turn*

### Open `07-causal-assumptions-exercises.qmd`

### Discuss with your neighbors: what do exchangeability, positivity, and consistency require for the question, does quitting smoking cause weight gain?

`r countdown::countdown(minutes = 10)`
