---
title: "Fitting the outcome model"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false
options(
  tibble.max_extra_cols = 6, 
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(causalworkshop)
library(marginaleffects)

set.seed(1234)
```

## Outcome Model {.large}

```{r}
#| eval: false
library(broom)

lm(outcome ~ exposure, data = df, weights = wts) |>
  tidy()
```

. . .

✅ This will get us the point estimate  

. . .

❌ This will NOT get us the correct confidence intervals

. . .

📦 There are three ways to get them right

## Three ways to estimate the uncertainty

::: {.small}
::: nonincremental
1. **A bootstrap**: correct and general purpose, but slow
2. **A robust standard error**: a sandwich estimator for the outcome model alone
3. **A propensity-aware sandwich**: correct and fast
:::
:::

. . .

The bootstrap is still worth knowing: it also solves problems that have no closed-form solution

## 1. Create a function to run your analysis once on a sample of your data

::: {.small}

```{r}
#| code-line-numbers: "|1-2|5-9|12-14|17-18"
fit_ipw <- function(.split, ...) {
  .df <- as.data.frame(.split)

  # fit propensity score model
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .df
  )

  # calculate inverse probability weights
  .df <- propensity_model |>
    augment(type.predict = "response", data = .df) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))

  # fit correctly bootstrapped ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```
:::

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| code-line-numbers: "|4-8"
#| output-location: slide
library(rsample)

# fit ipw model to bootstrapped samples
bootstrapped_net_data <- bootstraps(
  net_data,
  times = 1000,
  apparent = TRUE
)

bootstrapped_net_data
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
fit_ipw(bootstrapped_net_data$splits[[1]])
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| cache: true
#| output-location: slide
#| code-line-numbers: "|2"
ipw_results <- bootstrapped_net_data |>
  mutate(boot_fits = map(splits, fit_ipw)) 

ipw_results
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| echo: false
ipw_results |>
  # remove the original data set results
  filter(id != "Apparent") |>
  mutate(
    estimate = map_dbl(
      boot_fits,
      # pull the `estimate` for `netTRUE` for each fit
      \(.fit) .fit |>
        filter(term == "netTRUE") |>
        pull(estimate)
    )
  ) |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 30, fill = "#D55E00FF", color = "white", alpha = 0.8) +
  theme_minimal()
```


## 3. Pull out the causal effect

```{r}
#| code-line-numbers: "|2"
#| output-location: fragment
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, boot_fits) |>
  filter(term == "netTRUE")

boot_estimate
```

## Fit both models on the full data {.small}

```{r}
#| code-line-numbers: "|1-5|7-9|11"
propensity_model <- glm(
  net ~ income + health + temperature,
  family = binomial(),
  data = net_data
)

net_data_wts <- propensity_model |>
  augment(type.predict = "response", data = net_data) |>
  mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))

ipw_model <- lm(malaria_risk ~ net, data = net_data_wts, weights = wts)
```

## Robust standard errors are not enough {.small}

```{r}
#| code-line-numbers: "|1|4|5|6"
#| output-location: fragment
robust_se <- avg_comparisons(ipw_model, variables = "net", vcov = "HC3")

bind_rows(
  "lm() default" = tidy(ipw_model) |> filter(term == "netTRUE"),
  "HC3 robust SE" = tidy(robust_se),
  "ipw()" = tidy(ipw(propensity_model, ipw_model)),
  .id = "method"
) |>
  select(method, estimate, std.error)
```

. . .

❌ Robust SEs account for the weights, but not for estimating them

## ipw(): correct and fast

::: nonincremental
- `ipw()` uses a sandwich estimator that also accounts for the uncertainty in
  the propensity score model
- Bring your own models: you fit the propensity score and outcome models
  yourself, then supply them to `ipw()`
:::

. . .

```{r}
ipw_fit <- ipw(propensity_model, ipw_model)
```

## ipw(): correct and fast {.small}

```{r}
ipw_fit
```

. . .

`ipw()` reports the estimand along with the estimate

## The bootstrap and ipw() agree

```{r}
#| echo: false
bind_rows(
  "bootstrap" = boot_estimate |>
    select(estimate = .estimate, conf.low = .lower, conf.high = .upper),
  "ipw()" = ipw_fit |>
    tidy(conf.int = TRUE) |>
    select(estimate, conf.low, conf.high),
  .id = "method"
)
```

. . .

`ipw()` fits each model once and takes a fraction of a second; the bootstrap fits 1,001 pairs of models

## *Your Turn* {.small}

`r countdown::countdown(minutes = 10)`

### Create a function called `fit_ipw` that fits the propensity score model and the weighted outcome model for the effect between `park_extra_magic_morning` and `wait_minutes_posted_avg`

### Using the `bootstraps()` and `int_t()` functions to estimate the final effect.

### Fitting both models on the full data and using `ipw()` to compare the two intervals.

## Binary outcomes: an odds ratio {.small}

```{r}
#| code-line-numbers: "|1-2|9"
#| output-location: fragment
net_data_wts <- net_data_wts |>
  mutate(risk_over_50 = malaria_risk > 50)

glm(
  risk_over_50 ~ net,
  data = net_data_wts,
  weights = wts,
  # quasibinomial() avoids a warning about non-integer successes
  family = quasibinomial()
) |>
  tidy(exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == "netTRUE") |>
  select(estimate, conf.low, conf.high)
```

## Binary outcomes: a risk difference {.small}

```{r}
#| output-location: fragment
lm(risk_over_50 ~ net, data = net_data_wts, weights = wts) |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE") |>
  select(estimate, conf.low, conf.high)
```

. . .

The default standard errors still do not account for the propensity score model; the same bootstrap and sandwich approaches apply here
