---
title: "Fitting the outcome model"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(causalworkshop)
library(marginaleffects)

set.seed(1234)
```

## Outcome Model {.large}

```{r}
#| eval: false

library(broom)

lm(outcome ~ exposure, data = df, weights = wts) |>
  tidy()
```

. . .

✅ This will get us the point estimate

. . .

❌ This will NOT get us the correct confidence intervals

. . .

📦 There are three other approaches we will consider

## Three ways to estimate the uncertainty

::: {.small}
::::: nonincremental
1. **A bootstrap**: correct and general purpose, but slow
2. **A robust standard error**: a sandwich estimator for the outcome model alone
3. **A propensity-aware sandwich**: correct and fast
:::::
:::

. . .

The bootstrap is still worth knowing: it also solves problems that have no closed-form solution

## 1. Create a function to run your analysis once on a sample of your data

::: {.small}

```{r}
#| code-line-numbers: "|1-2|5-9|12-14|17-18"

fit_ipw <- function(.split, ...) {
  .df <- as.data.frame(.split)

  # fit propensity score model
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .df
  )

  # calculate inverse probability weights
  .df <- propensity_model |>
    augment(type.predict = "response", data = .df) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))

  # fit correctly bootstrapped ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```
:::

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| code-line-numbers: "|4-8"
#| output-location: slide

library(rsample)

# fit ipw model to bootstrapped samples
bootstrapped_net_data <- bootstraps(
  net_data,
  times = 1000,
  apparent = TRUE
)

bootstrapped_net_data
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
fit_ipw(bootstrapped_net_data$splits[[1]])
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| cache: true
#| output-location: slide
#| code-line-numbers: "|2"

ipw_results <- bootstrapped_net_data |>
  mutate(boot_fits = map(splits, fit_ipw))

ipw_results
```

## 2. Use {rsample} to bootstrap our causal effect

```{r}
#| echo: false

ipw_results |>
  # remove the original data set results
  filter(id != "Apparent") |>
  mutate(
    estimate = map_dbl(
      boot_fits,
      # pull the `estimate` for `netTRUE` for each fit
      \(.fit) .fit |>
        filter(term == "netTRUE") |>
        pull(estimate)
    )
  ) |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 30, fill = "#D55E00FF", color = "white", alpha = 0.8) +
  theme_minimal()
```

## 3. Pull out the causal effect

```{r}
#| code-line-numbers: "|2"
#| output-location: fragment

# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, boot_fits) |>
  filter(term == "netTRUE")

boot_estimate
```

## Fit both models on the full data {.small}

```{r}
#| code-line-numbers: "|1-5|7-9|11"

propensity_model <- glm(
  net ~ income + health + temperature,
  family = binomial(),
  data = net_data
)

net_data_wts <- propensity_model |>
  augment(type.predict = "response", data = net_data) |>
  mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))

ipw_model <- lm(malaria_risk ~ net, data = net_data_wts, weights = wts)
```

## Robust standard errors

The sandwich estimator gets its name from the form of the variance it calculates: a "meat" matrix built from the variability of each observation's residuals, sandwiched between two copies of a "bread" matrix that comes from the design of the model

. . .

When the model's assumptions hold, the bread and meat combine to give the usual model-based variance

. . .

When they do not, as with IPW weights, the sandwich form still gives a consistent variance estimate

. . .

That is why these are called "robust" standard errors

## Robust standard errors

```{r}
#| output-location: fragment

vcov(ipw_model)
```

. . .

`vcov()` returns the variance-covariance matrix that `tidy()` and `summary()` use under the hood; it treats the weights as fixed, which they are not

## Robust standard errors

```{r}
#| output-location: fragment

library(sandwich)

sandwich(ipw_model)
```

. . .

For the treatment coefficient, the sandwich variance is about 1.5 times the model-based one, so the model-based SE understates the uncertainty; for the intercept, it is smaller

## Robust standard errors

```{r}
#| output-location: fragment

robust_se <- avg_comparisons(ipw_model, variables = "net", vcov = "HC3")

robust_se
```

::: {.callout-note}
`avg_comparisons()` returns a marginal effect: the average of the predicted outcome difference across the observed covariate distribution. `vcov = "HC3"` requests the HC3 variant of the sandwich. We will see more of marginaleffects when we get to g-computation.
:::

. . .

❌ Robust SEs account for the weights, but not for estimating them

## ipw(): correct and fast

::: nonincremental
- `ipw()` uses a sandwich estimator that also accounts for the uncertainty in
  the propensity score model
- Bring your own models: you fit the propensity score and outcome models
  yourself, then supply them to `ipw()`
:::

. . .

```{r}
ipw_fit <- ipw(propensity_model, ipw_model)
```

## ipw(): correct and fast {.small}

```{r}
ipw_fit
```

. . .

`ipw()` reports the estimand along with the estimate

## Marginal and conditional

The marginal reading reports the population-averaged causal contrast: what would happen if everyone used a net versus if no one did

. . .

The conditional reading reports the outcome model's coefficients, with standard errors from the same jointly estimated sandwich

. . .

Both are computed either way; `as_conditional()` and `as_marginal()` change which one the result presents

## The conditional reading

```{r}
#| output-location: fragment

tidy(as_conditional(ipw_fit), conf.int = TRUE) |>
  select(term, estimate, std.error, conf.low, conf.high)
```

. . .

`netTRUE` matches the marginal `diff` here because the outcome model is linear in the exposure with an identity link

## Comparing the three approaches

```{r}
#| echo: false

bind_rows(
  "bootstrap" = boot_estimate |>
    select(estimate = .estimate, conf.low = .lower, conf.high = .upper),
  "HC3 robust SE" = robust_se |>
    tidy() |>
    select(estimate, conf.low, conf.high),
  "ipw()" = ipw_fit |>
    tidy(conf.int = TRUE) |>
    select(estimate, conf.low, conf.high),
  .id = "method"
)
```

. . .

`ipw()` fits each model once and takes a fraction of a second; the bootstrap fits 1,001 pairs of models

## *Your Turn*

`r countdown::countdown(minutes = 10)`

### Create a function called `fit_ipw` that fits the propensity score model and the weighted outcome model for the effect between `park_extra_magic_morning` and `wait_minutes_posted_avg`

### Using the `bootstraps()` and `int_t()` functions to estimate the final effect.

### Fitting both models on the full data and using `ipw()` to compare the two intervals, then `as_conditional()` to switch readings.

## Binary outcomes

```{r}
#| code-line-numbers: "|1-2|4-11|9"

net_data_wts <- net_data_wts |>
  mutate(risk_over_50 = malaria_risk > 50)

binary_outcome_model <- glm(
  risk_over_50 ~ net,
  data = net_data_wts,
  weights = wts,
  # quasibinomial() avoids a warning about non-integer successes
  family = quasibinomial()
)
```

## Binary outcomes: all three effects

```{r}
#| output-location: fragment

ipw(propensity_model, binary_outcome_model) |>
  tidy(conf.int = TRUE, exponentiate = TRUE) |>
  select(term, estimate, conf.low, conf.high)
```

. . .

`ipw()` reports the risk difference, risk ratio, and odds ratio, with standard errors that account for the propensity score model

. . .

`as_conditional()` returns the model's log-odds coefficient, which exponentiated is the `or` row; only the marginal reading has the risk difference and risk ratio
