---
title: "Continuous exposures and g-computation"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)

library(tidyverse)
library(broom)
library(causaldata)
library(marginaleffects)

set.seed(1000)
```

## {.huge .center background-color="#23373B"}

Normal regression estimates associations. But we want *causal* estimates: what would happen if *everyone* in the study were exposed to x vs if *no one* was exposed.

## G-Computation/G-Formula {background-color="#23373B"}

1. Fit a model for `y ~ x + z` where z is all covariates
2. Create a duplicate of your data set for each level of `x`
3. Set the value of x to a single value for each cloned data set (e.g `x = 1` for one, `x = 0` for the other)

## G-Computation/G-Formula {background-color="#23373B"}

4. Make predictions using the model on the cloned data sets
5. Calculate the estimate you want, e.g. `mean(x_1) - mean(x_0)`

## *Advantages of the parametric G-formula*

. . .

**Often more statistically precise than propensity-based methods**

. . .

**Incredibly flexible**

. . .

**Basis of other important causal models, e.g. causal survival analysis and TMLE**

## Greek Pantheon data (`greek_data`)

```{r}
#| echo: false

library(gt)
gt(head(greek_data, 10))
```

\+ 10 more rows

## 1. Fit a model for `y ~ a + l`

```{r}
greek_model <- lm(y ~ a + l, data = greek_data)
```

## 2. Create a duplicate of your data set for each level of `a`

```{r}
#| echo: false

gt(head(greek_data, 10))
```

## 2. Create a duplicate of your data set for each level of `a`

::: {.columns}

::::: {.column width="50%"}
```{r}
#| echo: false

gt(head(greek_data, 10))
```
:::::

::::: {.column width="50%"}
```{r}
#| echo: false

gt(head(greek_data, 10))
```
:::::

:::

## 3. Set the value of `a` to a single value for each cloned data set

::: {.columns}

::::: {.column width="50%"}
```{r}
#| echo: false

gt(head(greek_data |> mutate(a = 0), 10)) |>
  tab_style(
    style = list(
      cell_fill(color = "#CEE9FF"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(columns = a)
  )
```
:::::

::::: {.column width="50%"}
```{r}
#| echo: false

gt(head(greek_data |> mutate(a = 1), 10)) |>
  tab_style(
    style = list(
      cell_fill(color = "#CEE9FF"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(columns = a)
  )
```
:::::

:::

## 3. Set the value of `a` to a single value for each cloned data set

```{r}
#| code-line-numbers: "|3,7"

#  set all participants to have a = 0
untreated_data <- greek_data |>
  mutate(a = 0)

#  set all participants to have a = 1
treated_data <- greek_data |>
  mutate(a = 1)
```

## 4. Make predictions using the model on the cloned data sets

```{r}
#| code-line-numbers: "|2-3,7-8"

#  predict under the data where everyone is untreated
predicted_untreated <- greek_model |>
  augment(newdata = untreated_data) |>
  select(untreated = .fitted)

#  predict under the data where everyone is treated
predicted_treated <- greek_model |>
  augment(newdata = treated_data) |>
  select(treated = .fitted)

predictions <- bind_cols(
  predicted_untreated,
  predicted_treated
)
```

--------------------------------------------------------------------------------

## 5. Calculate the estimate you want

```{r}
#| eval: false
#| code-line-numbers: "|5"
#| output-location: fragment

predictions |>
  summarise(
    mean_treated = mean(treated),
    mean_untreated = mean(untreated),
    difference = mean_treated - mean_untreated
  )
```

```{r}
#| echo: false

predictions |>
  summarise(
    mean_treated = mean(treated),
    mean_untreated = mean(untreated),
    difference = mean_treated - mean_untreated
  ) |>
  mutate(across(everything(), \(x) round(x, digits = 1)))
```

## `avg_comparisons()` does all five steps

```{r}
#| eval: false

library(marginaleffects)

avg_comparisons(greek_model, variables = "a")
```

```{r}
#| echo: false

avg_comparisons(greek_model, variables = "a") |>
  tidy() |>
  select(estimate, std.error, conf.low, conf.high) |>
  mutate(across(everything(), \(x) round(x, digits = 2)))
```

The standard error comes from the **delta method**, which approximates the variance of a transformation of the estimated coefficients

## Targeting estimands with `newdata`

`newdata` changes whose covariates we average over

```{r}
#| eval: false
#| code-line-numbers: "|5,12"

# ATT: average over the treated
avg_comparisons(
  greek_model,
  variables = "a",
  newdata = filter(greek_data, a == 1)
)

# ATC: average over the untreated
avg_comparisons(
  greek_model,
  variables = "a",
  newdata = filter(greek_data, a == 0)
)
```

## Same model, three populations

```{r}
#| echo: false

bind_rows(
  ATE = tidy(avg_comparisons(greek_model, variables = "a")),
  ATT = tidy(avg_comparisons(
    greek_model,
    variables = "a",
    newdata = filter(greek_data, a == 1)
  )),
  ATC = tidy(avg_comparisons(
    greek_model,
    variables = "a",
    newdata = filter(greek_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Without an exposure-covariate interaction, the three coincide: the model assumes a constant effect across covariate values

## Tilting toward a target population

The ATO and ATM populations are not subsets of the rows, so weight the average by the tilting function instead

```{r}
#| code-line-numbers: "|3-4,9"

library(propensity)

greek_ps <- glm(a ~ l, data = greek_data, family = binomial()) |>
  predict(type = "response")

avg_comparisons(
  greek_model,
  variables = "a",
  wts = ps_tilt(greek_ps, "ato")
)
```

## {.huge .center background-color="#23373B"}

**Continuous exposures**

## {.huge .center background-color="#23373B"}

**We recommend g-computation over propensity scores for continuous exposures because of stability issues**

## Does change in smoking intensity (`smkintensity82_71`) affect weight gain among lighter smokers?

```{r}
nhefs_light_smokers <- nhefs_complete |>
  filter(smokeintensity <= 25)
```

## 1. Fit a model for `y ~ x + z`

```{r}
#| code-line-numbers: "|2|3-6"

nhefs_gcomp_model <- lm(
  wt82_71 ~ splines::ns(smkintensity82_71, df = 3) +
    sex +
    race +
    age +
    I(age^2) +
    education +
    smokeintensity +
    I(smokeintensity^2) +
    smokeyrs +
    I(smokeyrs^2) +
    exercise +
    active +
    wt71 +
    I(wt71^2),
  data = nhefs_light_smokers
)
```

G-computation still uses a single outcome model, and a continuous exposure lets us fit it flexibly with a natural spline. The remaining steps are identical, except that the levels of step 2 become the exposure values we choose to compare

## 2 and 3. Set `smkintensity82_71` in each cloned data set

```{r}
#| code-line-numbers: "|3,7"

#  set all participants to have no change in intensity
unchanged_data <- nhefs_light_smokers |>
  mutate(smkintensity82_71 = 0)

#  set all participants to smoke 10 fewer cigarettes
reduced_data <- nhefs_light_smokers |>
  mutate(smkintensity82_71 = -10)
```

## 4. Make predictions using the model on the cloned data sets

```{r}
#| code-line-numbers: "|2-3,7-8"

#  predict under the data where intensity is unchanged
predicted_unchanged <- nhefs_gcomp_model |>
  augment(newdata = unchanged_data) |>
  select(unchanged = .fitted)

#  predict under the data where everyone reduces by 10
predicted_reduced <- nhefs_gcomp_model |>
  augment(newdata = reduced_data) |>
  select(reduced = .fitted)

nhefs_predictions <- bind_cols(
  predicted_unchanged,
  predicted_reduced
)
```

## 5. Calculate the estimate you want

```{r}
#| code-line-numbers: "|5"
#| output-location: fragment

nhefs_predictions |>
  summarise(
    mean_unchanged = mean(unchanged),
    mean_reduced = mean(reduced),
    difference = mean_reduced - mean_unchanged
  )
```

## `avg_comparisons()` with chosen exposure values

```{r}
#| eval: false

avg_comparisons(
  nhefs_gcomp_model,
  variables = list(smkintensity82_71 = c(-10, 0))
)
```

```{r}
#| echo: false

avg_comparisons(
  nhefs_gcomp_model,
  variables = list(smkintensity82_71 = c(-10, 0))
) |>
  tidy() |>
  select(estimate, std.error, conf.low, conf.high) |>
  mutate(across(everything(), \(x) round(x, digits = 2)))
```

`avg_comparisons()` contrasts the higher value against the lower one, so the estimate is the mean at 0 minus the mean at -10. Reversing it gives the answer from step 5: reducing intensity by 10 cigarettes a day increases weight gain by 1.44 kg

## Predict across the exposure range

```{r}
nhefs_dose_response <- avg_predictions(
  nhefs_gcomp_model,
  variables = list(smkintensity82_71 = seq(-25, 25, by = 5))
)
```

With a continuous exposure, we don't have to settle for a single contrast: we can ask what the average weight gain would be at every value of the exposure

## Predict across the exposure range {.small}

```{r}
nhefs_dose_response
```

## The causal dose-response function

```{r}
#| echo: false

ggplot(nhefs_dose_response, aes(smkintensity82_71, estimate)) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill = "#E69F0095"
  ) +
  geom_line(linewidth = 1, color = "#E69F00") +
  theme_minimal(base_size = 20) +
  labs(
    x = "Change in smoking intensity",
    y = "Weight gain (kg)"
  )
```

## Do *posted* wait times at 8 am affect *actual* wait times at 9 am?

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5

library(tidyverse)
library(ggdag)
library(ggokabeito)

geom_dag_label_repel <- function(..., seed = 10) {
  ggdag::geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 1.25,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    family = "sans",
    seed = seed,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3,
    ...
  )
}

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather + extra,
  x ~ weather + close + Season + extra,
  extra ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(
    curvature = c(rep(0, 7), .2, 0, .2, .2, 0),
    edge_colour = "grey70"
  ) +
  geom_dag_point() +
  geom_dag_label_repel(seed = 1602) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-2.25, 2.25),
    breaks = c(-2, -1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "8am-9am\n(Today)",
      "9am-10am\n(Today)"
    )
  )
```

## *Your Turn*

### Work through Your Turns 1-4 in `11-continuous-g-computation-exercises.qmd`

`r countdown::countdown(minutes = 10)`
