---
title: "Continuous exposures and g-computation"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false
options(
  tibble.max_extra_cols = 6, 
  tibble.width = 60
)

library(tidyverse)
library(broom)
library(causaldata)
library(marginaleffects)

set.seed(1000)
```

## {background-color="#23373B" .huge .center}

Normal regression estimates associations. But we want *causal* estimates: what would happen if *everyone* in the study were exposed to x vs if *no one* was exposed.

## G-Computation/G-Formula {background-color="#23373B"}

1. Fit a model for `y ~ x + z` where z is all covariates
2. Create a duplicate of your data set for each level of `x` 
3. Set the value of x to a single value for each cloned data set (e.g `x = 1` for one, `x = 0` for the other)

## G-Computation/G-Formula {background-color="#23373B"}

4. Make predictions using the model on the cloned data sets
5. Calculate the estimate you want, e.g. `mean(x_1) - mean(x_0)`

## *Advantages of the parametric G-formula*

. . .

**Often more statistically precise than propensity-based methods**

. . .


**Incredibly flexible**

. . .

**Basis of other important causal models, e.g. causal survival analysis and TMLE**

## Greek Pantheon data (`greek_data`)

```{r}
#| echo: false
library(gt)
gt(head(greek_data, 10))
```

\+ 10 more rows

## 1. Fit a model for `y ~ a + l`

```{r}
greek_model <- lm(y ~ a + l, data = greek_data)
```

## 2. Create a duplicate of your data set for each level of `a` 

```{r}
#| echo: false
gt(head(greek_data, 10))
```

## 2. Create a duplicate of your data set for each level of `a` 

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
gt(head(greek_data, 10))
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
gt(head(greek_data, 10))
```
:::

::::

## 3. Set the value of `a` to a single value for each cloned data set

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: false
gt(head(greek_data |> mutate(a = 0), 10)) |>
  tab_style(
    style = list(
      cell_fill(color = "#CEE9FF"),
      cell_text(weight = "bold")
      ),
    locations = cells_body(columns = a)
  )
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
gt(head(greek_data |> mutate(a = 1), 10)) |>
  tab_style(
    style = list(
      cell_fill(color = "#CEE9FF"),
      cell_text(weight = "bold")
      ),
    locations = cells_body(columns = a)
  )
```
:::

::::

## 3. Set the value of `a` to a single value for each cloned data set

```{r}
#| code-line-numbers: "|3,7"
#  set all participants to have a = 0
untreated_data <- greek_data |>
  mutate(a = 0) 

#  set all participants to have a = 1
treated_data <- greek_data |>
  mutate(a = 1) 
```

## 4. Make predictions using the model on the cloned data sets

```{r}
#| code-line-numbers: "|2-3,7-8"
#  predict under the data where everyone is untreated
predicted_untreated <- greek_model |>
  augment(newdata = untreated_data) |>
  select(untreated = .fitted)

#  predict under the data where everyone is treated
predicted_treated <- greek_model |>
  augment(newdata = treated_data) |>
  select(treated = .fitted)

predictions <- bind_cols(
  predicted_untreated, 
  predicted_treated
) 
```

---

## 5. Calculate the estimate you want

```{r}
#| eval: false
#| code-line-numbers: "|5"
#| output-location: fragment
predictions |>
  summarise(
    mean_treated = mean(treated),
    mean_untreated = mean(untreated),
    difference = mean_treated - mean_untreated
  )
```

```{r}
#| echo: false
predictions |>
  summarise(
    mean_treated = mean(treated),
    mean_untreated = mean(untreated),
    difference = mean_treated - mean_untreated
  ) |>
  mutate(across(everything(), \(x) round(x, digits = 1)))
```

## Marginalizing over the covariates

```{r}
#| echo: false
bind_cols(greek_data, predictions) |>
  pivot_longer(
    c(untreated, treated),
    names_to = "scenario",
    values_to = "predicted"
  ) |>
  ggplot(aes(x = factor(l), y = predicted, color = scenario)) +
  geom_hline(
    yintercept = mean(predictions$treated),
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.25,
      dodge.width = 0.6,
      seed = 10
    ),
    size = 3,
    alpha = 0.8
  ) +
  annotate(
    "label",
    x = 1.5,
    y = mean(predictions$treated),
    label = "average prediction"
  ) +
  labs(x = "l", y = "predicted y", color = NULL)
```

**Predict at every observation's actual covariates, then average**

## `avg_comparisons()` does all five steps

```{r}
#| eval: false
library(marginaleffects)

avg_comparisons(greek_model, variables = "a")
```

```{r}
#| echo: false
avg_comparisons(greek_model, variables = "a") |>
  tidy() |>
  select(estimate, std.error, conf.low, conf.high) |>
  mutate(across(everything(), \(x) round(x, digits = 2)))
```

The standard error comes from the **delta method**, which approximates the variance of a transformation of the estimated coefficients

## Targeting estimands with `newdata`

`newdata` changes whose covariates we average over

```{r}
#| eval: false
#| code-line-numbers: "|5,12"
# ATT: average over the treated
avg_comparisons(
  greek_model,
  variables = "a",
  newdata = filter(greek_data, a == 1)
)

# ATC: average over the untreated
avg_comparisons(
  greek_model,
  variables = "a",
  newdata = filter(greek_data, a == 0)
)
```

## Same model, three populations

```{r}
#| echo: false
bind_rows(
  ATE = tidy(avg_comparisons(greek_model, variables = "a")),
  ATT = tidy(avg_comparisons(
    greek_model,
    variables = "a",
    newdata = filter(greek_data, a == 1)
  )),
  ATC = tidy(avg_comparisons(
    greek_model,
    variables = "a",
    newdata = filter(greek_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Without an exposure-covariate interaction, the three coincide: the model assumes a constant effect across covariate values

## Tilting toward a target population

The ATO and ATM populations are not subsets of the rows, so weight the average by the tilting function instead

```{r}
#| eval: false
library(propensity)

greek_ps <- glm(a ~ l, data = greek_data, family = binomial()) |>
  predict(type = "response")

avg_comparisons(
  greek_model,
  variables = "a",
  wts = ps_tilt(greek_ps, "ato")
)
```

## {background-color="#23373B" .huge .center} 
**Continuous exposures**

## {background-color="#23373B" .huge .center} 

**We recommend g-computation over propensity scores for continuous exposures because of stability issues**

## {background-color="#23373B" .center}

When overlap is poor, IPW makes its reliance on overlap explicit: the weights blow up and we are forced to confront the problem.

. . .

G-computation absorbs the problem silently, producing a stable estimate that may nonetheless be wrong.

## Do *posted* wait times at 8 am affect *actual* wait times at 9 am?

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
library(tidyverse)
library(ggdag)
library(ggokabeito)

geom_dag_label_repel <- function(..., seed = 10) {
  ggdag::geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 1.25, 
    inherit.aes = FALSE,
    max.overlaps = Inf, 
    family = "sans",
    seed = seed,
    label.size = NA, 
    label.padding = 0.1,
    size = 14 / 3,
    ...
  ) 
}

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
    y ~ x + close + Season + weather + extra,
    x ~ weather + close + Season + extra,
    extra ~ weather + close + Season,
    coords = coord_dag,
    labels = labels,
    exposure = "x",
    outcome = "y"
) |>
    tidy_dagitty() |>
    node_status() |>
    ggplot(
        aes(x, y, xend = xend, yend = yend, color = status)
    ) +
    geom_dag_edges_arc(curvature = c(rep(0, 7), .2, 0, .2, .2, 0), edge_colour = "grey70") +
    geom_dag_point() +
    geom_dag_label_repel(seed = 1602) + 
    scale_color_okabe_ito(na.value = "grey90") +
    theme_dag() +
    theme(
        legend.position = "none",
        axis.text.x = element_text()
    ) +
    coord_cartesian(clip = "off") +
    scale_x_continuous(
        limits = c(-2.25, 2.25),
        breaks = c(-2, -1, 0, 1, 2),
        labels = c(
            "\n(one year ago)",
            "\n(6 months ago)",
            "\n(3 months ago)",
            "8am-9am\n(Today)",
            "9am-10am\n(Today)"
        )
    )
```


## *Your Turn*

### Work through Your Turns 1-4 in `11-continuous-g-computation-exercises.qmd`

`r countdown::countdown(minutes = 10)`


