---
title: "Interaction and Heterogeneous Treatment Effects"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(marginaleffects)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- The ATE, ATT, and ATC are all expectations of the same unit-level contrast,
  `y(1) - y(0)`, taken over different covariate distributions
- That distinction becomes meaningful only when the individual treatment effect
  varies across units

## Two claims from earlier modules

- When standard methods succeed: the effect was homogeneous, so marginal and
  conditional effects agree
- In g-computation: without an exposure-covariate interaction, the ATE, ATT, and
  ATC coincide

. . .

This module is where both claims stop holding

## What makes the estimands differ

Any difference between them is driven by two things

- Variation in treatment effects across covariate profiles
- Differences in the distribution of those profiles across treated and untreated
  units

## {.center background-color="#23373B"}

### **Interaction terms parameterize heterogeneity**

## A constant treatment effect

- A model like `y ~ a + l` says the conditional mean is `b0 + b1 * a + b2 * l`
- The contrast between `a = 1` and `a = 0` is `b1`, whatever the value of `l`

. . .

The treatment effect is assumed to be the same for every unit, regardless of
covariate values

## Letting the effect vary

- Suppose one variable in the confounder set modifies the treatment effect, and
  call it `z`
- `y ~ a * z + l` adds the product term:
  `b0 + b1 * a + b2 * z + b3 * a * z + b4 * l`

. . .

The conditional treatment effect becomes `b1 + b3 * z`, which depends on `z`

## When is the effect constant?

- If `b3` is zero, the treatment effect is constant
- Otherwise the treatment effect changes across values of `z`

. . .

The interaction term formalizes the claim that the effect of treatment differs
across levels of a covariate

## Interaction or separate models?

- An equivalent representation is to fit separate models within strata of `z`,
  which lets the treatment coefficient differ across groups
- The interaction model is often preferred because it estimates those
  subgroup-specific effects within a single framework, using all the data

## {.center background-color="#23373B"}

### **A simulation where we know the answer**

## Simulate the potential outcomes

```{r}
#| code-line-numbers: "|5|6|7-8|9|10"

set.seed(1000)
n <- 10000

sim_data <- tibble(
  l = rnorm(n),
  z = factor(rbinom(n, 1, 0.5), labels = c("low", "high")),
  y0 = 10 + 2 * l + 3 * (z == "high") + rnorm(n),
  y1 = y0 + 2 + 4 * (z == "high"),
  a = rbinom(n, 1, plogis(-1.25 + 0.5 * l + 2.5 * (z == "high"))),
  y = if_else(a == 1, y1, y0)
)
```

## What the simulation says

- `l` is a continuous confounder: it affects both the exposure and the outcome
- `z` is a confounder too, and it also modifies the effect: the individual
  effect is 2 in the `low` stratum and 6 in the `high` stratum
- Both potential outcomes are simulated, so the answer is exact in this sample

## Who gets treated

```{r}
#| output-location: fragment

sim_data |>
  count(a, z) |>
  mutate(prop = n / sum(n), .by = a)
```

. . .

About three quarters of the treated are `high`, and about three quarters of the
untreated are `low`

## The truth in this simulation

```{r}
#| output-location: fragment
#| code-line-numbers: "|2|4-6"

sim_data |>
  mutate(effect = y1 - y0) |>
  summarize(
    ate = mean(effect),
    att = mean(effect[a == 1]),
    atc = mean(effect[a == 0])
  )
```

. . .

The ATT is larger than the ATC because the treated are mostly `high`, where the
effect is 6

## {.center background-color="#23373B"}

### **G-computation with an interaction**

## Fit the interaction model

```{r}
#| output-location: fragment

gcomp_model <- lm(y ~ a * z + l, data = sim_data)

tidy(gcomp_model)
```

## The coefficients are conditional

- `a` is the effect in the `low` stratum, and `a:zhigh` is how much larger the
  effect is in the `high` stratum
- Together they say the effect is about 2 and about 6

. . .

That is the conditional story, and averaging it over a population by eye is
another matter

## `avg_comparisons()` for the ATE

```{r}
#| output-location: fragment

avg_comparisons(gcomp_model, variables = "a")
```

## Targeting the ATT and the ATC

```{r}
#| eval: false
#| code-line-numbers: "|5,12"

# ATT: average over the exposed observations
avg_comparisons(
  gcomp_model,
  variables = "a",
  newdata = filter(sim_data, a == 1)
)

# ATC: average over the unexposed observations
avg_comparisons(
  gcomp_model,
  variables = "a",
  newdata = filter(sim_data, a == 0)
)
```

## Same model, three populations

```{r}
#| echo: false

bind_rows(
  ate = tidy(avg_comparisons(gcomp_model, variables = "a")),
  att = tidy(avg_comparisons(
    gcomp_model,
    variables = "a",
    newdata = filter(sim_data, a == 1)
  )),
  atc = tidy(avg_comparisons(
    gcomp_model,
    variables = "a",
    newdata = filter(sim_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Each estimate matches its simulated truth

## Averaging one response surface

Once interaction is introduced, these estimands differ because they average the
same heterogeneous response surface over different covariate distributions

## Effects within levels of `z`

```{r}
#| output-location: fragment

avg_comparisons(gcomp_model, variables = "a", by = "z")
```

## Testing the difference

```{r}
#| output-location: fragment
#| code-line-numbers: "|5"

avg_comparisons(
  gcomp_model,
  variables = "a",
  by = "z",
  hypothesis = ~pairwise
)
```

. . .

The two strata differ by about 4, with a standard error of about 0.05

## The effect in each stratum

```{r}
#| output-location: slide

plot_comparisons(gcomp_model, variables = "a", by = "z") +
  expand_limits(y = 0) +
  labs(x = "modifier", y = "effect of a on y")
```

## Specify every interaction

Because g-computation focuses on modeling the *outcome*, we must specify every
interaction we think exists in order to estimate treatment effect heterogeneity

## Drop the interaction

```{r}
gcomp_model_add <- lm(y ~ a + z + l, data = sim_data)
```

```{r}
#| echo: false

bind_rows(
  ate = tidy(avg_comparisons(gcomp_model_add, variables = "a")),
  att = tidy(avg_comparisons(
    gcomp_model_add,
    variables = "a",
    newdata = filter(sim_data, a == 1)
  )),
  atc = tidy(avg_comparisons(
    gcomp_model_add,
    variables = "a",
    newdata = filter(sim_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Without an exposure-covariate interaction, the three coincide: the model assumes
a constant effect across covariate values

## {.center background-color="#23373B"}

### **The same estimands through weights**

## IPW targets the same estimands

- IPW does not require an explicit regression model for the outcome surface
- The weighting scheme determines which covariate distribution the heterogeneous
  unit-level effects are averaged over

. . .

IPW captures heterogeneity implicitly through the weighting and averaging
target, not by estimating subgroup-specific effects

## Fit the propensity score model

```{r}
#| code-line-numbers: "|1|3-4|6-8"

propensity_model <- glm(a ~ l + z, family = binomial(), data = sim_data)

sim_data_wts <- propensity_model |>
  augment(type.predict = "response", data = sim_data) |>
  mutate(
    w_ate = wt_ate(.fitted, a),
    w_att = wt_att(.fitted, a),
    w_atu = wt_atu(.fitted, a)
  )
```

The modifier is a confounder, so it belongs in the propensity score model

## Three weighted outcome models

```{r}
outcome_ate <- lm(y ~ a, data = sim_data_wts, weights = w_ate)
outcome_att <- lm(y ~ a, data = sim_data_wts, weights = w_att)
outcome_atu <- lm(y ~ a, data = sim_data_wts, weights = w_atu)
```

- Each model regresses the outcome on the exposure alone, and only the weights
  change
- `wt_atu()` builds the weights for the unexposed observations, the ATU or ATC

## Estimates through `ipw()`

```{r}
#| output-location: fragment

ipw(propensity_model, outcome_ate) |>
  tidy(conf.int = TRUE) |>
  select(term, estimate, std.error, conf.low, conf.high)
```

## The three weighted estimates

```{r}
#| echo: false

bind_rows(
  ate = ipw(propensity_model, outcome_ate) |> tidy(conf.int = TRUE),
  att = ipw(propensity_model, outcome_att) |> tidy(conf.int = TRUE),
  atc = ipw(propensity_model, outcome_atu) |> tidy(conf.int = TRUE),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

The weighted estimates recover the same truths as g-computation

## The weights carry the estimand

- The weighted outcome model has no covariates, so the differences among the
  three estimates flow entirely through the weights
- The modifier separates the exposure groups strongly here, so the spread
  between the ATT and the ATC is visible

## {.center background-color="#23373B"}

### **Effect modification with `ipw()`**

## The subgroup question with weights

```{r}
#| output-location: fragment
#| code-line-numbers: "|1|3"

outcome_model_int <- lm(y ~ a * z, data = sim_data_wts, weights = w_ate)

ipw(propensity_model, outcome_model_int, .by = z) |>
  tidy(conf.int = TRUE) |>
  select(group, estimate, std.error, conf.low, conf.high)
```

## One system, one covariance

- The result reports the overall effect, one effect per level of `z`, and each
  level contrasted against the first
- Every row is a parameter of one jointly estimated system, so the stratum
  contrast has a standard error and is testable

. . .

Fitting each stratum on its own subset would report the same stratum effects
with no covariance between them, leaving the difference untestable

## What `.by` requires

- The modifier must be a factor or a character vector with no missing values
- Every level must hold both exposure levels, since a stratum in which nobody
  took one of them identifies no contrast against it

The remedy for an empty cell is a coarser modifier rather than a refit

## The model decides what varies

If the outcome model has no term reading the exposure and the modifier together,
the stratum effects are forced equal up to the covariate distribution of each
stratum, and the fit warns

. . .

The modification you report is then a property of the model rather than of the
data

## No odds ratios in the strata

The stratum rows report differences, and for a binary outcome risk differences
and log risk ratios, but never odds ratios

. . .

An odds ratio is noncollapsible, so a difference of two of them moves with the
outcome distribution whether or not the effect does

## {.center background-color="#23373B"}

### **Stabilizing the weights**

## Stabilized weights, revisited

- `stabilize = TRUE` puts the marginal probability of the observed exposure in
  the numerator, as it did when we first built ATE weights
- Under effect modification, the numerator should condition on the modifier
  instead

## Condition the numerator on `z`

```{r}
#| code-line-numbers: "|1-2|6|11"

numerator_model <- glm(a ~ z, family = binomial(), data = sim_data)
p_z <- fitted(numerator_model)

sim_data_wts <- sim_data_wts |>
  mutate(
    sw_ate = wt_ate(.fitted, a, stabilize = TRUE),
    sw_ate_z = wt_ate(
      .fitted,
      a,
      stabilize = TRUE,
      stabilization_score = if_else(a == 1, p_z, 1 - p_z)
    )
  )
```

`fitted()` alone gives the probability of exposure for every unit, which is not
the probability of the exposure an unexposed unit took

## The rule for the numerator

- The numerator may condition on any variable the outcome model already reads,
  without changing what is estimated
- Conditioning on a variable the model does not read changes the estimand: the
  pseudo-population it builds is one where that variable still predicts exposure

. . .

Condition the numerator on the modifier and on nothing the model does not hold

## The weights tighten

```{r}
#| output-location: fragment

sim_data_wts |>
  select(w_ate, sw_ate, sw_ate_z) |>
  # the three weight types stack only as plain numbers
  mutate(across(everything(), as.numeric)) |>
  pivot_longer(everything(), names_to = "weights") |>
  summarize(sd = sd(value), max = max(value), .by = weights)
```

. . .

The exposure is nearly balanced, so the default numerator is close to a constant
one half; conditioning on `z` is what reshapes the weights

## The same fit, stabilized

```{r}
#| output-location: fragment

outcome_model_stab <- lm(y ~ a * z, data = sim_data_wts, weights = sw_ate_z)

ipw(propensity_model, outcome_model_stab, .by = z) |>
  tidy(conf.int = TRUE) |>
  select(group, estimate, std.error, conf.low, conf.high)
```

## What stabilizing changed here

- The estimates and their standard errors are unchanged: the numerator is
  constant within each exposure-by-modifier cell, and the reported effects are
  averages within those cells
- The weights themselves are far less variable, which is where stabilizing pays

## One caveat on the variance

A supplied stabilization score is treated as known in the variance, while the
default stabilizer is estimated within the stacked system

## {.center background-color="#23373B"}

### **Two jobs for interaction terms**

## Interaction in the exposure model

- In the diagnostics module we added `income * health` to the propensity score
  model because that product term was imbalanced
- That interaction serves balance, and through balance exchangeability

. . .

It makes no claim about whether the effect varies

## Interaction in the outcome model

- The interaction in this module's outcome model encodes how the effect itself
  varies across levels of the modifier
- Same syntax, different jobs: which one you need depends on which model reads
  the term

## {.center background-color="#23373B"}

### **Effect modification versus causal interaction**
