---
title: "Bonus: Colliders, selection bias, and loss to follow-up"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)

library(tidyverse)

set.seed(100111)
```

## Confounders and chains

```{r}
#| echo: false

library(ggdag)
ggdag_confounder_triangle(x_y_associated = TRUE) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Colliders

```{r}
#| echo: false

ggdag_collider_triangle() +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Colliders

```{r}
#| echo: false

ggdag_adjust(
  collider_triangle(),
  "m",
  collider_lines = FALSE
) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## Let's prove it!

```{r}
set.seed(8)
collider_data <- collider_triangle() |>
  simulate_data(-0.6)
```

## Let's prove it!

```{r}
collider_data
```

## Let's prove it!

```{r}
#| echo: false

fit_model <- function(fmla) {
  lm(fmla, data = collider_data) |>
    broom::tidy(conf.int = TRUE) |>
    dplyr::filter(term == "x") |>
    dplyr::mutate(formula = as.character(fmla)[3]) |>
    dplyr::select(formula, estimate, conf.low, conf.high)
}

bind_rows(
  fit_model(y ~ x),
  fit_model(y ~ x + m)
) |>
  ggplot(aes(x = estimate, y = formula, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linewidth = 1, color = "grey80") +
  geom_pointrange(color = "steelblue", size = 0.75) +
  theme_minimal(18) +
  labs(
    y = NULL,
    caption = "correct effect size: 0"
  )
```

## Loss to follow-up

```{r}
#| echo: false

l2fu <- dagify(
  follow_up ~ symptoms,
  symptoms ~ new_rx + dx_severity,
  cd4 ~ dx_severity,
  labels = c(
    follow_up = "Follow-Up",
    symptoms = "Symptoms",
    new_rx = "New HIV Drug",
    dx_severity = "Underlying \nHIV Severity",
    cd4 = "CD4 Count"
  ),
  exposure = "new_rx",
  outcome = "cd4"
)

l2fu |>
  ggdag_adjust(
    "follow_up",
    layout = "mds",
    use_text = FALSE,
    collider_lines = FALSE
  ) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 100,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot() +
  ggokabeito::scale_color_okabe_ito()
```

## Censoring as missingness

```{r}
#| echo: false

cd4_labels <- c(
  cd4_missing = "Missingness \nin CD4 Count",
  symptoms = "Symptoms",
  new_rx = "New HIV Drug",
  dx_severity = "Underlying \nHIV Severity",
  cd4 = "CD4 Count",
  unknown = "Unknown \nMechanism"
)

missing_cd4 <- dagify(
  cd4_missing ~ symptoms,
  symptoms ~ new_rx + dx_severity,
  cd4 ~ dx_severity,
  coords = list(
    x = c(
      new_rx = 0,
      symptoms = 1.2,
      cd4_missing = 0,
      dx_severity = 3,
      cd4 = 4.6
    ),
    y = c(new_rx = 2, symptoms = 1, cd4_missing = 0, dx_severity = 1, cd4 = 1)
  ),
  labels = cd4_labels,
  exposure = "new_rx",
  outcome = "cd4"
)

missing_cd4 |>
  ggdag_adjust("cd4_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## When censoring doesn't bias

```{r}
#| echo: false

missing_cd4_benign <- dagify(
  cd4_missing ~ new_rx + unknown,
  symptoms ~ new_rx + dx_severity,
  cd4 ~ dx_severity,
  coords = list(
    x = c(
      new_rx = 0,
      symptoms = 1.2,
      cd4_missing = 0,
      dx_severity = 3,
      cd4 = 4.6,
      unknown = -1.4
    ),
    y = c(
      new_rx = 2,
      symptoms = 1,
      cd4_missing = 0,
      dx_severity = 1,
      cd4 = 1,
      unknown = 1
    )
  ),
  labels = cd4_labels,
  exposure = "new_rx",
  outcome = "cd4"
)

missing_cd4_benign |>
  ggdag_adjust("cd4_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## What can we recover? {.small}

::: nonincremental
- We are always conditioning on the data we actually have
- When that opens no backdoor path, complete-case analysis still recovers the causal effect, with less precision
- We may still be unable to estimate other statistics, like the mean of the outcome
- When it does open a backdoor path, sometimes we can close it and sometimes we can't
- Not all censoring needs weights
:::

## Change the estimand {.small}

::: nonincremental
- Sometimes the variable you select on is both a collider and a mediator
- Conditioning on it induces relationships between other variables, and it blocks the indirect effect
- We can't calculate the total effect, but we can still calculate the direct effect by adjusting for the selection variable
:::

## Adjusting for selection bias {background-color="#23373B"}

1. Fit a probability of censoring model, e.g. *glm(censoring ~ predictors, family = binomial())*
2. Create weights using inverse probability strategy
3. Use weights in your causal model

## {.large .center background-color="#23373B"}

### We won't do it here, but you can include many types of weights in a given model. Just take their product, e.g. *multiply inverse propensity of treatment weights by inverse propensity of censoring weights*.

## *Your Turn*

### Work through Your Turns 1-3 in `16-bonus-selection-bias-exercises.qmd`

`r countdown::countdown(minutes = 10)`
