---
title: "Continuous exposures with propensity scores"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)

library(tidyverse)
library(broom)
library(causaldata)
library(touringplans)
library(propensity)
library(marginaleffects)
library(halfmoon)
library(positively)
library(balancing)

set.seed(1000)
```

## {.huge .center background-color="#23373B"}

### *Warning!* Propensity score weights are sensitive to positivity violations for continuous exposures.

## {.huge .center background-color="#23373B"}

### **The story so far**

---

## Propensity score weighting {background-color="#23373B"}

1. Fit a propensity model predicting exposure `x`, `x + z` where z is all covariates
2. Calculate weights
3. Fit an outcome model estimating the effect of `x` on `y` weighted by the propensity score

## Continous exposures {background-color="#23373B"}

1. Use a model like `lm(x ~ z)` for the propensity score model.
2. Use `wt_ate()` with `.fitted`; transforms using `dnorm()` to get on probability-like scale.
3. Apply the weights to the outcome model as normal!

## Alternative: quantile binning {.small background-color="#23373B"}

1. Bin the continuous exposure into quantiles and use categorical regression like a multinomial model to calculate probabilities.
2. Calculate the weights where the propensity score is the probability you fall into the quantile you actually fell into. Same as the binary ATE!
3. Same workflow for the outcome model

## 1. Fit a model for `exposure ~ confounders`

```{r}
#| eval: false

model <- lm(
  exposure ~ confounder_1 + confounder_2,
  data = df
)
```

## 2. Calculate the weights with `wt_ate()`

```{r}
#| eval: false
#| code-line-numbers: "|3-8"

model |>
  augment(data = df) |>
  # wt_ate() estimates the SD from the model
  mutate(wts = wt_ate(
    .fitted,
    exposure,
    exposure_type = "continuous"
  ))
```

## Does change in smoking intensity (`smkintensity82_71`) affect weight gain among lighter smokers?

```{r}
nhefs_light_smokers <- nhefs_complete |>
  filter(smokeintensity <= 25)
```

## 1. Fit a model for `exposure ~ confounders`

```{r}
#| code-line-numbers: "|1-2|3-6"

nhefs_model <- lm(
  smkintensity82_71 ~ sex + race + age + I(age^2) +
    education + smokeintensity + I(smokeintensity^2) +
    smokeyrs + I(smokeyrs^2) + exercise + active +
    wt71 + I(wt71^2),
  data = nhefs_light_smokers
)
```

## 2. Calculate the weights with `wt_ate()`

```{r}
#| code-line-numbers: "|3-7"

nhefs_wts <- nhefs_model |>
  augment(data = nhefs_light_smokers) |>
  mutate(wts = wt_ate(
    .fitted,
    smkintensity82_71,
    exposure_type = "continuous"
  ))
```

## 2. Calculate the weights with `wt_ate()`

```{r}
nhefs_wts
```

## Do *posted* wait times at 8 am affect *actual* wait times at 9 am?

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6.5

library(tidyverse)
library(ggdag)
library(ggokabeito)

geom_dag_label_repel <- function(..., seed = 10) {
  ggdag::geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 1.25,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    family = "sans",
    seed = seed,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3,
    ...
  )
}

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather + extra,
  x ~ weather + close + Season + extra,
  extra ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(
    curvature = c(rep(0, 7), 0.2, 0, 0.2, 0.2, 0),
    edge_colour = "grey70"
  ) +
  geom_dag_point() +
  geom_dag_label_repel(seed = 1602) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-2.25, 2.25),
    breaks = c(-2, -1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "8am-9am\n(Today)",
      "9am-10am\n(Today)"
    )
  )
```

## *Your Turn 1*

### Fit a model using `lm()` with `wait_minutes_posted_avg` as the outcome and the confounders identified in the DAG.

### Use `augment()` to add model predictions to the data frame

### In `wt_ate()`, calculate the weights using `.fitted` and `wait_minutes_posted_avg`

`r countdown::countdown(minutes = 5)`

## *Your Turn 1*

```{r}
#| include: false

eight <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 8) |>
  select(-wait_minutes_actual_avg)

nine <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  select(park_date, wait_minutes_actual_avg)

wait_times <- eight |>
  left_join(nine, by = "park_date") |>
  drop_na(wait_minutes_actual_avg)
```

```{r}
post_time_model <- lm(
  wait_minutes_posted_avg ~
    park_close +
    park_extra_magic_morning +
    park_temperature_high +
    park_ticket_season,
  data = wait_times
)
```

## *Your Turn 1*

```{r}
wait_times_wts <- post_time_model |>
  augment(data = wait_times) |>
  mutate(
    wts = wt_ate(
      .fitted,
      wait_minutes_posted_avg,
      exposure_type = "continuous"
    )
  )
```

## *Stabilizing extreme weights*

```{r}
#| echo: false

nhefs_wts |>
  ggplot(aes(wts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") +
  scale_x_log10() +
  theme_minimal(base_size = 20) +
  xlab("Weights")
```

## Stabilizing extreme weights {background-color="#23373B"}

1. Fit an intercept-only model (e.g. `lm(x ~ 1)`) or use mean and SD of `x`
2. Calculate the marginal density from this model.
3. Divide the marginal density by the propensity score density. `wt_ate(.., stabilize = TRUE)` does this all!

## Calculate stabilized weights

```{r}
#| code-line-numbers: "|7"

nhefs_swts <- nhefs_model |>
  augment(data = nhefs_light_smokers) |>
  mutate(swts = wt_ate(
    .fitted,
    smkintensity82_71,
    exposure_type = "continuous",
    stabilize = TRUE
  ))
```

## Stabilizing extreme weights

```{r}
#| echo: false

ggplot(nhefs_swts, aes(swts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") +
  scale_x_log10() +
  theme_minimal(base_size = 20) +
  xlab("Stabilized Weights")
```

## Stabilized weights should have a mean close to 1

```{r}
mean(nhefs_swts$swts)
```

### If the mean is far from 1, we may have issues with model misspecification or positivity violations. A check worth running every time.

## *Your Turn 2*

### Re-fit the above using stabilized weights

`r countdown::countdown(minutes = 3)`

## *Your Turn 2*

```{r}
wait_times_swts <- post_time_model |>
  augment(data = wait_times) |>
  mutate(
    swts = wt_ate(
      .fitted,
      wait_minutes_posted_avg,
      exposure_type = "continuous",
      stabilize = TRUE
    )
  )
```

## Assessing positivity {background-color="#23373B"}

1. A continuous exposure has no groups to compare
2. Instead, we ask how much data supports each value we intend to set
3. `check_edp()` reports that as an effective number of observations

## Effective data points

```{r}
edp <- check_edp(
  nhefs_swts,
  smkintensity82_71,
  c(age, wt71, smokeintensity, smokeyrs),
  values = c(-20, 0, 20),
  exposure_type = "continuous"
)

edp |>
  tidy() |>
  group_by(value) |>
  summarize(edp = round(mean(edp), 1))
```

## Effective data points

- Setting everyone to a 20 cigarette increase is supported by about 6 effective observations out of 1162
- The dose-response function is weakest where the effective sample is smallest

## Checking balance {background-color="#23373B"}

1. Weighting should break the association between the confounders and the exposure
2. For a binary exposure, we check standardized mean differences
3. For a continuous exposure, we check correlations and energy

## Check balance with `check_balance()`

```{r}
balance_metrics <- check_balance(
  nhefs_swts,
  .vars = c(age, smokeintensity, smokeyrs, wt71),
  .exposure = smkintensity82_71,
  .weights = swts,
  exposure_type = "continuous"
)
```

### These are the continuous confounders; the categorical ones need dummy coding first

### As in `wt_ate()`, `exposure_type` tells `check_balance()` the exposure is continuous

## Check balance with `check_balance()`

```{r}
balance_metrics
```

### The weighted correlations are close to zero, and energy is smaller after weighting

## Correlations before and after weighting

```{r}
#| echo: false

balance_metrics |>
  filter(metric == "correlation") |>
  ggplot(aes(x = estimate, y = variable, group = method, color = method)) +
  geom_love(vline_xintercept = 0) +
  scale_color_okabe_ito() +
  theme_minimal(base_size = 20) +
  scale_x_continuous(breaks = c(-0.1, 0)) +
  labs(x = "Correlation", y = NULL, color = NULL)
```

## Fitting the outcome model {background-color="#23373B"}

1. Use the stabilized weights in the outcome model. Nothing new here!
2. Use `ipw()` for standard errors that account for the weights

## Fit the outcome model

```{r}
#| code-line-numbers: "|3"

nhefs_ipw_model <- lm(
  wt82_71 ~ smkintensity82_71,
  weights = swts,
  data = nhefs_swts
)
```

---

```{r}
#| code-line-numbers: "|4"

nhefs_ipw_model |>
  tidy() |>
  filter(term == "smkintensity82_71") |>
  mutate(estimate = estimate * -10)
```

## Estimate the effect with `ipw()`

```{r}
nhefs_ipw_fit <- ipw(nhefs_model, nhefs_ipw_model)
```

### `ipw()` uses a sandwich estimator that also accounts for the uncertainty in the propensity score model

## Estimate the effect with `ipw()` {.small}

```{r}
nhefs_ipw_fit
```

## *Your Turn 3*

### Estimate the relationship between posted wait times and actual wait times using the stabilized weights we just created, saving the model to an object

### Use `ipw()` to estimate the effect with standard errors that account for the weights, then `as_conditional()` to switch readings

`r countdown::countdown(minutes = 5)`

## *Your Turn 3*

```{r}
ipw_model <- lm(
  wait_minutes_actual_avg ~ wait_minutes_posted_avg,
  weights = swts,
  data = wait_times_swts
)

ipw_model |>
  tidy() |>
  filter(term == "wait_minutes_posted_avg") |>
  mutate(estimate = estimate * 10)
```

## *Your Turn 3*

```{r}
ipw_fit <- ipw(post_time_model, ipw_model)
```

### `ipw()` reports the effect of one more minute, so multiply by 10 to compare

## *Your Turn 3* {.small}

```{r}
ipw_fit
```

## *Your Turn 3*

```{r}
as_conditional(ipw_fit) |>
  tidy(conf.int = TRUE)
```

### The slope matches the marginal reading here because the outcome model is linear in the exposure with an identity link

## Balancing weights {background-color="#23373B"}

1. The stabilized weights leave small correlations behind
2. Balancing weights optimize those correlations directly
3. As with `wt_ate()`, we declare the exposure type

## Entropy balancing with `balance()`

```{r}
#| output-location: slide

entropy_fit <- balance(
  nhefs_swts,
  .exposure = smkintensity82_71,
  .covariates = c(age, wt71, smokeintensity, smokeyrs),
  method = bw_entropy(),
  exposure_type = "continuous"
)

entropy_fit
```

## Comparing balance

```{r}
nhefs_swts <- nhefs_swts |>
  mutate(w_bal = weights(entropy_fit))

balance_comparison <- check_balance(
  nhefs_swts,
  .vars = c(age, smokeintensity, smokeyrs, wt71),
  .exposure = smkintensity82_71,
  .weights = c(swts, w_bal),
  exposure_type = "continuous"
)
```

## Comparing balance {.small}

```{r}
# `group_level` is the exposure name on every row here
balance_comparison |>
  select(-group_level)
```

## Comparing balance

```{r}
#| echo: false
#| fig-width: 9

plot_balance(balance_comparison) +
  scale_x_continuous(n.breaks = 3) +
  ggokabeito::scale_color_okabe_ito() +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

## Estimating with balancing weights

```{r}
balancing_outcome_model <- lm(
  wt82_71 ~ smkintensity82_71,
  weights = w_bal,
  data = nhefs_swts
)

balancing_ipw_fit <- ipw(entropy_fit, balancing_outcome_model)
```

## Estimating with balancing weights {.small}

```{r}
balancing_ipw_fit
```

## Causal dose-response functions {background-color="#23373B"}

1. Calculate the stabilized weights
2. Fit the outcome model with a flexible term for the exposure
3. Use `ipw()` for standard errors that account for the weights
4. Predict the outcome across the range of the exposure

## 1. Calculate the stabilized weights

### `nhefs_swts` already has the stabilized weights

## 2. Fit a flexible outcome model

```{r}
nhefs_msm <- lm(
  wt82_71 ~ splines::ns(smkintensity82_71, df = 3),
  weights = swts,
  data = nhefs_swts
)
```

## 3. Estimate the effects with `ipw()`

```{r}
nhefs_ipw <- ipw(
  nhefs_model,
  nhefs_msm,
  .data = nhefs_swts
)
```

### The standard errors account for the estimated weights.

## 4. Predict across the exposure range {.small}

```{r}
#| output-location: slide

dose_response <- nhefs_ipw |>
  as_conditional() |>
  # marginaleffects needs the data to build the grid
  set_modeldata(nhefs_swts) |>
  avg_predictions(
    variables = list(smkintensity82_71 = -25:25)
  )

dose_response
```

### The predictions along the exposure grid come from the outcome model itself, so we switch to the conditional reading first

## The causal dose-response function

```{r}
#| echo: false

ggplot(dose_response, aes(smkintensity82_71, estimate)) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill = "#E69F0095"
  ) +
  geom_line(linewidth = 1, color = "#E69F00") +
  theme_minimal(base_size = 20) +
  labs(
    x = "Change in smoking intensity",
    y = "Weight gain (kg)"
  )
```

## Dose-response with balancing weights

```{r}
#| echo: false

balancing_msm <- lm(
  wt82_71 ~ splines::ns(smkintensity82_71, df = 3),
  weights = w_bal,
  data = nhefs_swts
)

dose_response_bal <- ipw(entropy_fit, balancing_msm) |>
  as_conditional() |>
  set_modeldata(nhefs_swts) |>
  avg_predictions(
    variables = list(smkintensity82_71 = -25:25)
  )

ggplot(dose_response_bal, aes(smkintensity82_71, estimate)) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill = "#E69F0095"
  ) +
  geom_line(linewidth = 1, color = "#E69F00") +
  theme_minimal(base_size = 20) +
  labs(
    x = "Change in smoking intensity",
    y = "Weight gain (kg)"
  )
```

## Diagnosing issues {background-color="#23373B"}

1. Extreme weights even after stabilization
2. Bootstrap: non-normal distribution
3. Bootstrap: estimate different from original model

## More info {background-color="#23373B"}

### [Causal Inference in R](https://www.r-causal.org/chapters/13-g-comp): g-computation for continuous exposures

### https://github.com/LucyMcGowan/writing-positivity-continous-ps
