---
title: "Causal Modeling in R: Whole Game"
format: html
---

```{r}
#| label: setup

options(tipr.verbose = FALSE, propensity.quiet = TRUE)
library(tidyverse)
library(broom)
library(rsample)
library(ggdag)
library(ggokabeito)
library(causalworkshop)
library(propensity)
library(halfmoon)
library(estimatr)
library(tipr)

set.seed(1234)
```

## Causal Modeling: Whole Game

In this guided exercise, we'll attempt to answer a causal question: does
mosquito bed net use reduce the risk of malaria? Bed nets prevent malaria by
putting a barrier between sleeping people and the mosquitoes that carry the
parasite, and several randomized trials in the 1990s established that they work.
There are still good reasons to ask the question with observational data: a
trial may be too slow, too expensive, or, now that we know nets are effective,
unethical to run.

To answer this question, we'll draw our assumptions with a causal diagram (a
directed acyclic graph, or DAG), which will guide our model. Then we'll use
inverse probability weighting, one of many causal modeling techniques, to
estimate the causal effect we're interested in. Finally, we'll ask how much
unmeasured confounding it would take to change our conclusion.

We'll use simulated data, `net_data`, from the {causalworkshop} package, which
includes ten variables:

`id`
:   an ID variable

`net`
:   a logical variable that is TRUE if the participant used a net and FALSE if
    they didn't

`net_num`
:   the same variable coded numerically, 1 if the participant used a net and 0
    if they didn't

`malaria_risk`
:   a risk of malaria scale ranging from 0-100

`income`
:   weekly income, measured in dollars

`health`
:   a health score scale ranging from 0-100

`household`
:   number of people living in the household

`eligible`
:   a binary variable indicating if the household is eligible for the free net
    program

`temperature`
:   the average temperature at night, in Celsius

`insecticide_resistance`
:   insecticide resistance of local mosquitoes, measured on a scale of 0-100,
    with higher values indicating higher resistance

Our thanks to Andrew Heiss for providing these data!

```{r}
library(causalworkshop)
net_data
```

Let's look at the distribution of malaria risk between the two groups.

```{r}
net_data |>
  ggplot(aes(malaria_risk, fill = net)) +
  geom_density(color = NA, alpha = 0.8) +
  scale_fill_okabe_ito(order = c(1, 5)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "malaria risk (0-100)",
    fill = "used a net"
  )
```

The density of those who used nets sits to the left of those who did not, so net
use looks protective. Let's look at the numeric summaries.

```{r}
net_data |>
  group_by(net) |>
  summarize(
    mean_malaria_risk = mean(malaria_risk),
    sd = sd(malaria_risk),
    .groups = "drop"
  )
```

Households that used a net had a mean malaria risk of 27.5, compared with 43.9
among households that did not, a difference of about 16 points. But is there
something else that could explain these results? Many factors are associated
with both net use and malaria risk; could one of those factors explain away what
we're seeing here?

To answer that, we need to specify a causal diagram based on domain knowledge.
For most circumstances, there is no data-driven approach that consistently
identifies confounders. Only our causal assumptions can help us identify them.
Causal diagrams are a visual expression of those assumptions linked to rigorous
mathematics that let us work out what we need to account for in our model.

In R, we can visualize and analyze our DAGs with the {ggdag} package. {ggdag}
uses {ggplot2} and {ggraph} to visualize diagrams and {dagitty} to analyze them.
The `dagify()` function takes formulas, much like `lm()` and friends, to express
assumptions. Here we say that malaria risk is caused by net use, income, health,
nighttime temperatures, and insecticide resistance; that net use is caused by
income, health, temperature, eligibility for the free net program, and household
size; that eligibility is determined by income and household size; and that
health is caused by income. We also identify the exposure and the outcome, lay
the nodes out by hand, and add labels so the diagram is easier to read.

```{r}
mosquito_dag <- dagify(
  malaria_risk ~ net + income + health + temperature + resistance,
  net ~ income + health + temperature + eligible + household,
  eligible ~ income + household,
  health ~ income,
  exposure = "net",
  outcome = "malaria_risk",
  coords = list(
    x = c(
      malaria_risk = 7,
      net = 3,
      income = 4,
      health = 5,
      temperature = 6,
      resistance = 8.5,
      eligible = 2,
      household = 1
    ),
    y = c(
      malaria_risk = 2,
      net = 2,
      income = 3,
      health = 1,
      temperature = 3,
      resistance = 2,
      eligible = 3,
      household = 2
    )
  ),
  labels = c(
    malaria_risk = "Risk of malaria",
    net = "Mosquito net",
    income = "Income",
    health = "Health",
    temperature = "Nighttime temperatures",
    resistance = "Insecticide resistance",
    eligible = "Eligible for program",
    household = "Number in the household"
  )
) |>
  tidy_dagitty()

mosquito_dag
```

Let's visualize our assumptions.

```{r}
mosquito_dag |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 3.5,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    seed = 10,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3
  ) +
  theme_dag(base_size = 14) +
  theme(legend.position = "none") +
  labs(caption = "Thanks to Andrew Heiss for the data!") +
  coord_cartesian(clip = "off") +
  scale_color_okabe_ito(na.value = "grey90")
```

If this DAG is right, the association we saw above is a mixture of eight open
paths: the causal path from net use to malaria risk and seven confounding paths.
What do we need to control for to block those seven? In many DAGs there will be
several sets of variables, called adjustment sets, that do the job.
`ggdag_adjustment_set()` visualizes them. Here there is only one: income,
health, and temperature.

```{r}
mosquito_dag |>
  ggdag_adjustment_set(use_text = FALSE, use_labels = TRUE, shadow = TRUE) +
  theme_dag() +
  scale_color_okabe_ito(order = c(1, 5)) +
  scale_fill_okabe_ito(order = c(1, 5))
```

Let's fit a model with these variables. Because the data are simulated, we know
we don't need splines or interactions here; in practice, you often do.

```{r}
lm(
  malaria_risk ~ net + income + health + temperature,
  data = net_data
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")
```

Adjusting for the variables in our adjustment set gives an estimate of about
-12.0 (95% CI -12.6, -11.4), a good deal smaller than the unadjusted difference
of 16 points. However, we are trying to answer a specific causal question: how
much would malaria risk change if a person used a net compared with if that same
person did not? Let's use an inverse probability weighting model to estimate
that effect at the population level (what if *everyone* used a net versus what
if *no one* did).

For a simple IPW model, we have two modeling steps. First, we fit a propensity
score model, which predicts the probability of the exposure (here, that a
household used a net). We use this model to calculate inverse probability
weights, the inverse of the probability of the exposure you actually received.
Then, in the second step, we use those weights in the outcome model, which
estimates the effect of the exposure on the outcome.

For the propensity score model, we'll use logistic regression, since net use is
a binary variable. The outcome is net use, and the variables in the model are
those in our adjustment set. Then we'll use `augment()` from {broom}, which
calls `predict()` on the inside, to calculate our weights with
`propensity::wt_ate()` and save them back into our data set.

```{r}
propensity_model <- glm(
  net ~ income + health + temperature,
  family = binomial(),
  data = net_data
)

net_data_wts <- propensity_model |>
  # predict whether a household used a net
  augment(data = net_data, type.predict = "response") |>
  # calculate inverse probability
  mutate(wts = wt_ate(.fitted, net))

net_data_wts |>
  select(net, .fitted, wts)
```

Let's look at the distribution of the weights.

```{r}
ggplot(net_data_wts, aes(wts)) +
  geom_histogram(color = "white", fill = "#E69F00", bins = 50) +
  theme_minimal(base_size = 20) +
  xlab("Weights")
```

The weights are skewed, but the largest is under 9, so there are no outrageous
values. If we saw extreme weights, we might try trimming or stabilizing them, or
consider a different estimand.

The main goal here is to *break* the non-causal associations between net use and
malaria risk, the other paths that might distort our results. In other words, if
we succeed, there should be no differences in the confounders between our two
groups. This is where randomized trials shine; you can often assume that there
are no baseline differences among potential confounders between treatment
groups.

Standardized mean differences (SMDs) are a simple measurement of differences
that work across variable types. In general, the closer to 0 we are, the better
job we have done eliminating the non-causal relationships we drew in our DAG.
Note that low SMDs for everything we adjust for does *not* mean that there is
nothing else confounding our study. Unmeasured confounders or misspecified DAGs
can still distort our effects, even if our SMDs look great.

We'll use the {halfmoon} package to calculate the SMDs, then visualize them.

```{r}
plot_df <- tidy_smd(
  net_data_wts,
  c(income, health, temperature),
  .group = net,
  .wts = wts
)

ggplot(
  plot_df,
  aes(x = abs(smd), y = variable, group = method, color = method)
) +
  geom_love()
```

A standard guideline is that balanced confounders should have an SMD of less
than 0.1 on the absolute scale. Income and health are well above that line
before weighting; after weighting, all three confounders sit close to 0.

We can also use halfmoon's `geom_mirror_histogram()` to visualize the impact the
weights are having on our population.

```{r}
ggplot(net_data_wts, aes(.fitted)) +
  geom_mirror_histogram(
    aes(group = net),
    bins = 50
  ) +
  geom_mirror_histogram(
    aes(fill = net, weight = wts),
    bins = 50,
    alpha = 0.5
  ) +
  scale_y_continuous(labels = abs) +
  labs(x = "propensity score") +
  theme_minimal(base_size = 20)
```

Both groups are being *upweighted* so that their distributions of propensity
scores are much more similar.

We could do more here to analyze our assumptions, but let's move on to our
second step: fitting the outcome model weighted by our inverse probabilities.
Some researchers call these marginal structural models, in part because the
model is marginal; we only need to include our outcome (`malaria_risk`) and
exposure (`net`). The other variables aren't in the model; they are accounted
for by the weights.

```{r}
ipw_model <- lm(
  malaria_risk ~ net,
  data = net_data_wts,
  weights = wts # inverse probability weights
)

ipw_estimate <- ipw_model |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")

ipw_estimate
```

This estimate, -12.5, is close to what we saw from the multivariable regression.
For simple causal questions this is often the case: adjusting for confounders
directly in the regression model sometimes estimates the same effect as IPW and
other causal techniques. Causal techniques are special, though, because they use
counterfactual modeling, which lets us handle circumstances such as selection
bias and time-dependent confounding.

But we have another problem to address. While we're using `lm()` to estimate our
IPW model, it doesn't properly account for the fact that we estimated the
weights. That means our standard error is too small, which artificially narrows
confidence intervals and artificially shrinks p-values. There are many ways to
address this, including robust estimators. We'll focus on the bootstrap via the
{rsample} package, but here's one way to do it with robust standard errors:

```{r}
# also see robustbase, survey, gee, and others
ipw_model_robust <- lm_robust(
  malaria_risk ~ net,
  data = net_data_wts,
  weights = wts
)

ipw_estimate_robust <- ipw_model_robust |>
  tidy(conf.int = TRUE) |>
  filter(term == "netTRUE")

as_tibble(ipw_estimate_robust)
```

Now let's try the bootstrap. First, we need to wrap our model in a function so
we can call it many times on our bootstrapped data. A function like this might
be your instinct; however, it's not quite right.

```{r}
# fit ipw model for a single bootstrap sample
fit_ipw_not_quite_rightly <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # fit ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```

The problem is that we need to account for the *entire* modeling process, so we
need to include the first step of our analysis, fitting the inverse probability
weights.

```{r}
fit_ipw <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # fit propensity score model
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .df
  )

  # calculate inverse probability weights
  .df <- propensity_model |>
    augment(type.predict = "response", data = .df) |>
    mutate(wts = wt_ate(.fitted, net))

  # fit correctly bootstrapped ipw model
  lm(malaria_risk ~ net, data = .df, weights = wts) |>
    tidy()
}
```

{rsample} makes the rest easy for us: `bootstraps()` resamples our data 1000
times, then we can use `purrr::map()` to apply our function to each resampled
set (`splits`). {rsample}'s `int_*()` functions help us get confidence intervals
for our estimate.

```{r}
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(net_data, times = 1000, apparent = TRUE) |>
  mutate(results = map(splits, fit_ipw))

# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) |>
  filter(term == "netTRUE")

boot_estimate
```

Let's compare the three sets of confidence intervals.

```{r}
bind_rows(
  ipw_estimate |>
    select(estimate, conf.low, conf.high) |>
    mutate(type = "ols"),
  ipw_estimate_robust |>
    select(estimate, conf.low, conf.high) |>
    mutate(type = "robust"),
  boot_estimate |>
    select(estimate = .estimate, conf.low = .lower, conf.high = .upper) |>
    mutate(type = "bootstrap")
) |>
  #  calculate CI width to sort by it
  mutate(width = conf.high - conf.low) |>
  arrange(width) |>
  #  fix the order of the model types for the plot
  mutate(type = fct_inorder(type)) |>
  ggplot(aes(x = type, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "#0172B1", size = 0.75) +
  coord_flip() +
  theme_minimal(base_size = 20) +
  theme(axis.title.y = element_blank())
```

So we have an estimate for our causal effect. The average treatment effect of
net use on malaria risk, comparing a world where everyone used a net with one
where no one did, is -12.5 (95% CI -13.4, -11.6).

# Sensitivity analysis

Getting the right answer to a causal question depends on getting our assumptions
more or less right. But what if we're more on the less side?

Spoiler alert: the answer we just calculated is wrong.

When conducting a causal analysis, it's a good idea to use sensitivity analyses
to test your assumptions. There are many potential sources of bias in a study
and many sensitivity analyses to go with them; here we'll focus on the
assumption of no unmeasured confounding.

Let's start broad. When we have little information about an unmeasured
confounder, we can use a tipping point analysis to ask how much confounding it
would take to tip our estimate to the null. The {tipr} package is a toolkit for
these analyses. `tip_coef()` takes an estimate, here the upper bound of our
confidence interval, which is closer to 0, and either the scaled difference in
means of the confounder between exposure groups or the effect of the confounder
on the outcome. We'll ask for five scenarios, where the scaled mean difference
in the confounder between exposure groups is 1, 2, 3, 4, or 5.

```{r}
tipping_points <- tip_coef(
  boot_estimate$.upper,
  exposure_confounder_effect = 1:5
)

tipping_points |>
  select(exposure_confounder_effect, confounder_outcome_effect)
```

```{r}
tipping_points |>
  ggplot(aes(confounder_outcome_effect, exposure_confounder_effect)) +
  geom_line(color = "#009E73", linewidth = 1.1) +
  geom_point(fill = "#009E73", color = "white", size = 2.5, shape = 21) +
  theme_minimal(base_size = 20) +
  labs(
    x = "Confounder-Outcome Effect",
    y = "Scaled mean difference in\nconfounder between\nexposure groups"
  )
```

If we had an unmeasured confounder whose standardized mean difference between
exposure groups was 1, it would need to reduce malaria risk by about 11.6 to tip
our upper confidence bound to 0. That is strong relative to the other effects in
this analysis. If instead the relationship between net use and the unmeasured
confounder were very strong, with a scaled mean difference of 5, the
confounder-malaria relationship would only need to be -2.3. Which of these
scenarios is plausible given our domain knowledge?

Now let's consider a much more specific sensitivity analysis. Some ethnic groups
have a genetic resistance to malaria. Let's say that in our simulated data, an
unnamed group in the unnamed country shares this resistance, and that for
historical reasons net use in this group is also very high. We don't have this
variable in `net_data`, but let's say we know from the literature that:

1. People with this genetic resistance have, on average, a malaria risk lower by
   about 10.
2. About 26% of people who use nets in our study have this genetic resistance.
3. About 5% of people who don't use nets have this genetic resistance.

With this much information, we can use {tipr} to adjust the estimates we
calculated for the unmeasured confounder. `adjust_coef_with_binary()` calculates
the adjusted estimates.

```{r}
adjusted_estimates <- boot_estimate |>
  select(.estimate, .lower, .upper) |>
  unlist() |>
  adjust_coef_with_binary(
    exposed_confounder_prev = 0.26,
    unexposed_confounder_prev = 0.05,
    confounder_outcome_effect = -10
  )

adjusted_estimates
```

The adjusted estimate for a situation where genetic resistance to malaria is a
confounder is -10.4 (95% CI -11.3, -9.5).

In fact, these data were simulated with just such a confounder. The true DAG
that generated the data is the one we drew plus one more variable, genetic
resistance to malaria, which reduces malaria risk and increases net use. It is
thus a confounder and part of the minimal adjustment set. By leaving it out, we
calculated the wrong effect.

```{r}
mosquito_dag_full <- dagify(
  malaria_risk ~ net + income + health + temperature +
    insecticide_resistance + genetic_resistance,
  net ~ income + health + temperature + eligible +
    household + genetic_resistance,
  eligible ~ income + household,
  health ~ income,
  exposure = "net",
  outcome = "malaria_risk",
  coords = list(
    x = c(
      malaria_risk = 7,
      net = 3,
      income = 4,
      health = 5,
      temperature = 6,
      insecticide_resistance = 8.5,
      eligible = 2,
      household = 1,
      genetic_resistance = 8.5
    ),
    y = c(
      malaria_risk = 2,
      net = 2,
      income = 3,
      health = 1,
      temperature = 3,
      insecticide_resistance = 2,
      eligible = 3,
      household = 2,
      genetic_resistance = 1
    )
  ),
  labels = c(
    malaria_risk = "Risk of malaria",
    net = "Mosquito net",
    income = "Income",
    health = "Health",
    temperature = "Nighttime temperatures",
    insecticide_resistance = "Insecticide resistance",
    eligible = "Eligible for program",
    household = "Number in the household",
    genetic_resistance = "Malaria resistance"
  )
) |>
  tidy_dagitty()

mosquito_dag_full |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 3.5,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    seed = 10,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3
  ) +
  theme_dag(base_size = 14) +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  scale_color_okabe_ito(na.value = "grey90")
```

The unmeasured confounder is available in `net_data_full` as
`genetic_resistance`. Let's add it to the propensity score model and recalculate
the IPW estimate.

```{r}
# add genetic resistance to the propensity score model
propensity_model_full <- glm(
  net ~ income + health + temperature + genetic_resistance,
  family = binomial(),
  data = net_data_full
)

net_data_full_wts <- propensity_model_full |>
  augment(type.predict = "response", data = net_data_full) |>
  mutate(wts = wt_ate(.fitted, net))

ipw_model_full <- lm(
  malaria_risk ~ net,
  data = net_data_full_wts,
  weights = wts
)

ipw_fit_full <- ipw(propensity_model_full, ipw_model_full)
tidy(ipw_fit_full, conf.int = TRUE)
```

With the confounder measured, the estimate is -10.2 (95% CI -11.2, -9.3).
Because these data are simulated, we know the true effect of net use on malaria
risk is about -10, so our corrected estimate is close.

What do you think? Is this estimate reliable? Did we do a good job addressing
the assumptions we need to make for a causal effect, particularly that there is
no confounding? How might you criticize this analysis, and what would you do
differently? In practice we can never check our answer against the truth, so we
need to keep probing our assumptions until we are confident they are robust.

# Take aways

- The broad strokes of a causal analysis are: 1) specify your causal question 2)
  make your assumptions clear 3) model and diagnose those assumptions 4) use the
  right estimator for the question you're asking and 5) ask how sensitive your
  answer is to the assumptions you can't check. As scientists, we should be able
  to critique each of these steps, and that's a good thing.
- To create marginal structural models, first fit a propensity score model for
  the weights with the exposure as the outcome. Then use the inverse of the
  predicted probabilities as weights in a model with just the outcome and
  exposure.
- No amount of modeling fixes an unmeasured confounder. Sensitivity analyses
  tell you how much unmeasured confounding it would take to change your
  conclusion.
