---
title: "Using Propensity Scores"
format: html
---

```{r}
#| label: setup

library(tidyverse)
library(broom)
library(touringplans)
library(propensity)
library(halfmoon)
library(positively)
```

We are interested in examining the relationship between whether there were "Extra Magic Hours" in the morning (the **exposure**) and the average posted wait time for the Seven Dwarfs Mine Train the same day between 9 AM and 10 AM in 2018 (the **outcome**).

Below is the propensity score model you created in the previous exercise.

```{r}
seven_dwarfs <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9)

propensity_model <- glm(
  park_extra_magic_morning ~ park_ticket_season +
    park_close +
    park_temperature_high,
  data = seven_dwarfs,
  family = binomial()
)

seven_dwarfs_prop <- propensity_model |>
  augment(type.predict = "response", data = seven_dwarfs)
```

## Your Turn 1 (Matching)

Create a "matched" data set using the same propensity score model as above and a caliper of 0.2.

1. Provide `matchit()` the formula for the propensity score.
2. Using the `link` and `caliper` arguments, create a caliper of 0.2 SDs on the linear logit scale.
3. Extract the matched data set into a new data frame called `matched_df`.

```{r}
library(MatchIt)
matched_dwarfs <- matchit(
  ___,
  data = seven_dwarfs,
  ___,
  ___
)

matched_df <- ___(matched_dwarfs)

matched_df
```

## Your Turn 2 (Weighting)

Add the ATE weights to the data frame, `seven_dwarfs_prop`

```{r}
seven_dwarfs_prop <- seven_dwarfs_prop |>
  mutate(w_ate = ___)
```

## Stretch goal: ATM weights

Add ATM weights to the data frame, `seven_dwarfs_prop`

```{r}
seven_dwarfs_prop <- seven_dwarfs_prop |>
  mutate(w_atm = ___)
```

## Stretch goal: The weighted distribution

Update the code below to examine the distribution of the weighted sample. **HINT**: update the `weight` parameter in the second `geom_mirror_histogram()` call using the ATE weights you created in Your Turn 2, then try the ATM weights.

```{r}
#| warning: false

seven_dwarfs_prop |>
  mutate(
    park_extra_magic_morning = factor(park_extra_magic_morning)
  ) |>
  ggplot(aes(.fitted)) +
  geom_mirror_histogram(
    aes(group = park_extra_magic_morning),
    bins = 30
  ) +
  geom_mirror_histogram(
    aes(fill = park_extra_magic_morning, weight = ___),
    alpha = 0.5,
    bins = 30
  ) +
  geom_hline(yintercept = 0, lwd = 0.5) +
  theme_minimal() +
  scale_y_continuous(labels = abs) +
  scale_fill_manual(values = c("blue", "green")) +
  labs(x = "p", fill = "Extra Magic Morning")
```

## Stretch goal: Low-support regions

Weights and trimming respond to extreme propensity scores.
`check_port()` looks for the covariate regions that produce them.
Fill in the confounders from the propensity score model above.

```{r}
# `park_close` is stored in seconds, which is hard to read in the cutpoints
seven_dwarfs_hours <- seven_dwarfs |>
  mutate(park_close = as.numeric(park_close) / 3600)

port <- check_port(
  seven_dwarfs_hours,
  park_extra_magic_morning,
  c(___, ___, ___)
)

port |>
  tidy() |>
  filter(low_support) |>
  select(description, n, prevalence)
```

One of these subgroups had no Extra Magic Mornings at all.
Is that a structural violation or a stochastic one?
