---
title: "Outcome Model"
format: html
---


```{r}
#| label: setup
library(tidyverse)
library(broom)
library(touringplans)
library(rsample)
library(propensity)

seven_dwarfs <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9)
```

We are interested in examining the relationship between whether there were "Extra Magic Hours" in the morning (the **exposure**) and the average wait time for the Seven Dwarfs Mine Train the same day between 9am and 10am (the **outcome**).

## Your turn

Create a function called `fit_ipw` that fits the propensity score model and incorporates the ATE weights (as calculated in the previous exercises), then uses the weights in a weighted outcome model.

```{r}
fit_ipw <- function(split, ...) { 
  .df <- ____
  
  # fit propensity score model
  
  # calculate ATE weights

  # fit correctly bootstrapped ipw model 
  lm(___ ~ ___, data = .df, weights = ___) |>
    tidy() 
}
```

Bootstrap this result 1000 times.

```{r}
set.seed(1234)

ipw_results <- ____(___, 1000, apparent = TRUE) |>
  mutate(boot_fits = map(splits, _____)) 
```

Check out the distribution of estimates (**no need to change this code**)

```{r}
ipw_results |>
  mutate(
    estimate = map_dbl(
      boot_fits,
      # pull the `estimate` for `park_extra_magic_morning` for each fit
      \(.fit) .fit |>
        filter(term == "park_extra_magic_morning") |>
        pull(estimate)
    )
  ) |>
  ggplot(aes(estimate)) +
  geom_histogram(fill = "#D55E00FF", color = "white", alpha = 0.8) + 
  theme_minimal()
```

Calculate the confidence interval

```{r}
boot_estimate <- ____(____, ____) |>
  filter(term == ____)

boot_estimate
```


Stretch goal: Do the same for a model using matching. Bootstrapping matched data the same way is not valid: matching induces a dependence between matched pairs, and resampling rows breaks it. There are two valid routes, and both start from a matched data set that we create once and then leave fixed.

```{r}
library(MatchIt)

matched_dwarfs <- matchit(
  park_extra_magic_morning ~ park_ticket_season +
    park_close +
    park_temperature_high,
  data = seven_dwarfs
)

matched_data <- get_matches(matched_dwarfs)

matched_outcome <- lm(___ ~ ___, data = matched_data)
```

The first route is a cluster-robust standard error, clustered within matched pairs. `get_matches()` returns a `subclass` column that identifies each pair; pass it to `vcovCL()` as a formula.

```{r}
library(sandwich)

vcl <- vcovCL(____, cluster = ~____)

tibble(
  estimate = coef(matched_outcome)[["park_extra_magic_morning"]],
  robust_se = sqrt(vcl["park_extra_magic_morning", "park_extra_magic_morning"]),
  conf.low = estimate - 1.96 * robust_se,
  conf.high = estimate + 1.96 * robust_se
)
```

The second route is a cluster bootstrap: resample whole matched pairs rather than rows, and do not re-run `matchit()` inside the bootstrap.

```{r}
fit_matched <- function(.split, ...) {
  .df <- as.data.frame(.split)
  lm(___ ~ ___, data = .df) |>
    tidy()
}

cluster_boots <- ____(
  matched_data,
  group = ____,
  times = 1000,
  apparent = TRUE
)

cluster_results <- cluster_boots |>
  mutate(boot_fits = map(splits, ____))

int_t(cluster_results, boot_fits) |>
  filter(term == ____)
```
