---
title: "Continuous exposures and g-computation"
format: html
---

```{r}
#| label: setup

library(tidyverse)
library(broom)
library(touringplans)
library(splines)
library(marginaleffects)
library(positively)
```

For this set of exercises, we'll use g-computation to calculate a causal effect for continuous exposures.

In the touringplans data set, we have information about the posted waiting times for rides. We also have a limited amount of data on the observed, actual times. The question that we will consider is this: Do posted wait times (`wait_minutes_posted_avg`) for the Seven Dwarfs Mine Train at 8 am affect actual wait times (`wait_minutes_actual_avg`) at 9 am? Here's our DAG:

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggdag)
library(ggokabeito)

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather + extra,
  x ~ weather + close + Season + extra,
  extra ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(
    curvature = c(rep(0, 7), 0.2, 0, 0.2, 0.2, 0),
    edge_colour = "grey70"
  ) +
  geom_dag_point() +
  geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 1.25,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    family = "sans",
    seed = 1602,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3
  ) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-2.25, 2.25),
    breaks = c(-2, -1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "8am-9am\n(Today)",
      "9am-10am\n(Today)"
    )
  )
```

First, let's wrangle our data to address our question: do posted wait times at 8 affect actual wait times at 9? We'll join the baseline data (all covariates and posted wait time at 8) with the outcome (average actual time). We also have a lot of missingness for `wait_minutes_actual_avg`, so we'll drop unobserved values for now.

You don't need to update any code here, so just run this.

```{r}
eight <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 8) |>
  select(-wait_minutes_actual_avg)

nine <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  select(park_date, wait_minutes_actual_avg)

wait_times <- eight |>
  left_join(nine, by = "park_date") |>
  drop_na(wait_minutes_actual_avg)
```

## Your Turn 1

For the parametric G-formula, we'll use a single model to fit a causal model of Posted Waiting Times (`wait_minutes_posted_avg`) on Actual Waiting Times (`wait_minutes_actual_avg`) where we  include all covariates, much as we normally fit regression models. However, instead of interpreting the coefficients, we'll calculate the estimate by predicting on cloned data sets.

Two additional differences in our model: we'll use a natural cubic spline on the exposure, `wait_minutes_posted_avg`, using `ns()` from the splines package, and we'll include an interaction term between `wait_minutes_posted_avg` and `park_extra_magic_morning`. These complicate the interpretation of the coefficient of the model in normal regression but have virtually no downside (as long as we have a reasonable sample size) in g-computation, because we still get an easily interpretable result. The interaction also decides which estimands are available to us: without one, the ATE, the ATT, and the ATC from g-computation are identical, because the model assumes the effect of the exposure is the same at every covariate value.

First, let's fit the model. 

1. Use `lm()` to create a model with the outcome, exposure, and confounders identified in the DAG. 
2. Save the model as `standardized_model`

```{r}
_______ ___ _______(
  wait_minutes_actual_avg ~ ns(_______, df = 2)*park_extra_magic_morning + _______ + _______ + _______, 
  data = wait_times
)
```

## Your Turn 2

Now that we've fit a model, we need to clone our data set. To do this, we'll simply mutate it so that in one set, all participants have `wait_minutes_posted_avg` set to 30 minutes and in another, all participants have `wait_minutes_posted_avg` set to 60 minutes. 

1. Create the cloned data sets, called `thirty` and `sixty`.
2. For both data sets, use `standardized_model` and `augment()` to get the predicted values. Use the `newdata` argument in `augment()` with the relevant cloned data set. Then, select only the fitted value. Rename `.fitted` to either `thirty_posted_minutes` or `sixty_posted_minutes` (use the pattern `select(new_name = old_name)`).
3. Save the predicted data sets as `predicted_thirty` and `predicted_sixty`.

```{r}
_______ <- wait_times |>
  _______

_______ <- wait_times |>
  _______

predicted_thirty <- standardized_model |>
  _______(newdata = _______) |>
  _______

predicted_sixty <- standardized_model |>
  _______(newdata = _______) |>
  _______
```

## Your Turn 3

Finally, we'll get the mean differences between the values. 

1. Bind `predicted_thirty` and  `predicted_sixty` using `bind_cols()`
2. Summarize the predicted values to create three new variables: `mean_thirty`, `mean_sixty`, and `difference`. The first two should be the means of `thirty_posted_minutes` and `sixty_posted_minutes`. `difference` should be `mean_sixty` minus `mean_thirty`.

```{r}
_______ |>
  _______(
    mean_thirty = _______,
    mean_sixty = _______,
    difference = _______
  )
```

That's it! `difference` is our effect estimate, marginalized over the spline terms, interaction effects, and confounders.

## Your Turn 4

Now let's calculate the same estimate the short way. `avg_comparisons()` from the marginaleffects package does every step you just did by hand, and it attaches a standard error and a confidence interval calculated with the delta method.

1. Pass `standardized_model` to `avg_comparisons()`.
2. Use the `variables` argument to set the two posted wait times you want to compare, 30 and 60 minutes.

```{r}
avg_comparisons(
  _______,
  variables = list(wait_minutes_posted_avg = c(___, ___))
)
```

The point estimate is the `difference` you calculated in Your Turn 3. The interval is the new information, and it is the lesson here: it is wide enough to include large effects in both directions, so this effect is not precisely estimated.

## Stretch goal: Bootstrapped intervals

The delta method is not the only route to an interval. As with propensity-based models, we can also bootstrap. In this stretch goal, use rsample to bootstrap the estimates we got from the G-computation model.

Remember, you need to bootstrap the entire modeling process, including the regression model, cloning the data sets, and calculating the effects.

```{r}
set.seed(1234)
library(rsample)

fit_gcomp <- function(.split, ...) { 
  # get bootstrapped data frame
  .df <- as.data.frame(.split)
  
  # fit outcome model. remember to model using `.df` instead of `wait_times`
  
  
  # clone datasets. remember to clone `.df` instead of `wait_times`
  
  
  # predict actual wait time for each cloned dataset
  
  
  # calculate ATE
  bind_cols(predicted_thirty, predicted_sixty) |>
    summarize(
      mean_thirty = mean(thirty_posted_minutes),
      mean_sixty = mean(sixty_posted_minutes),
      difference = mean_sixty - mean_thirty
    ) |>
    # rsample expects a `term` and `estimate` column
    pivot_longer(everything(), names_to = "term", values_to = "estimate")
}

gcomp_results <- bootstraps(wait_times, 1000, apparent = TRUE) |>
  mutate(results = map(splits, ______))

# using bias-corrected confidence intervals
boot_estimate <- int_bca(_______, results, .fn = fit_gcomp)

boot_estimate
```

## Stretch goal: The causal dose-response function

With a continuous exposure, we don't have to settle for a single contrast. We can ask what the average actual wait time would be at every posted wait time, which traces out a causal dose-response function. `avg_predictions()` marginalizes the same way `avg_comparisons()` does: at each value of the exposure, it predicts at every observation's own covariate values and averages those predictions.

1. Use `avg_predictions()` on `standardized_model`, setting the exposure to a grid of values with the `variables` argument.
2. Plot the estimates as a line with a confidence band, using `conf.low` and `conf.high`.

```{r}
dose_response <- avg_predictions(
  _______,
  variables = list(wait_minutes_posted_avg = seq(10, 80, by = 5))
)

dose_response |>
  ggplot(aes(x = wait_minutes_posted_avg, y = estimate)) +
  geom_ribbon(aes(ymin = _______, ymax = _______), alpha = 0.2) +
  geom_line(linewidth = 1) +
  labs(
    x = "Posted wait time at 8 am (minutes)",
    y = "Average actual wait at 9 am (minutes)"
  )
```

The grid runs from 10 to 80 minutes because that is about the range of posted wait times we observe; asking for predictions outside it would make the spline extrapolate. Notice how much of the plot the confidence band takes up, and compare that to the interval you got in Your Turn 4.

## Stretch goal: How much data supports each clone?

Staying inside the observed range does not mean the predictions are equally well supported across it. In Your Turn 2 you set every day's posted wait to 30 and then to 60. `check_edp()` reports the effective number of observations that support each of those interventions, given the covariates.

1. Set `values` to the two exposure values you cloned to.
2. `wait_minutes_posted_avg` is a continuous exposure, so declare the exposure type rather than letting `check_edp()` detect it.

```{r}
edp <- check_edp(
  wait_times,
  wait_minutes_posted_avg,
  c(
    park_close,
    park_extra_magic_morning,
    park_temperature_high,
    park_ticket_season
  ),
  values = c(___, ___),
  exposure_type = ___
)

edp |>
  tidy() |>
  group_by(value) |>
  summarize(edp = round(mean(edp), 1))
```

Both are a small fraction of the 85 days in the data, under two days of effective support each, and the 60 minute clone is the smaller of the two. That is the same shortage of support the wide interval in Your Turn 4 reports.

---

# Takeaways

- To fit the parametric G-formula, fit a standardized model with all covariates. Then, use cloned data sets with values set to each level of the exposure you want to study. 
- Use the model to predict the values for that level of the exposure and compute the effect estimate you want
- `avg_comparisons()` runs those steps in a single call and returns a standard error and confidence interval with the estimate
