---
title: "Propensity scores for continuous exposures"
format: html
---

```{r}
#| label: setup

library(tidyverse)
library(broom)
library(touringplans)
library(propensity)
```

For this set of exercises, we'll use propensity scores for continuous exposures.

In the touringplans data set, we have information about the posted waiting times for rides. We also have a limited amount of data on the observed, actual times. The question that we will consider is this: Do posted wait times (`wait_minutes_posted_avg`) for the Seven Dwarves Mine Train at 8 am affect actual wait times (`wait_minutes_actual_avg`) at 9 am? Here’s our DAG:

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggdag)
library(ggokabeito)

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather + extra,
  x ~ weather + close + Season + extra,
  extra ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(
    curvature = c(rep(0, 7), 0.2, 0, 0.2, 0.2, 0),
    edge_colour = "grey70"
  ) +
  geom_dag_point() +
  geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 1.25,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    family = "sans",
    seed = 1602,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3
  ) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-2.25, 2.25),
    breaks = c(-2, -1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "8am-9am\n(Today)",
      "9am-10am\n(Today)"
    )
  )
```

First, let’s wrangle our data to address our question: do posted wait times at 8 affect actual weight times at 9? We’ll join the baseline data (all covariates and posted wait time at 8) with the outcome (average actual time). We also have a lot of missingness for `wait_minutes_actual_avg`, so we’ll drop unobserved values for now.

You don't need to update any code here, so just run this.

```{r}
eight <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 8) |>
  select(-wait_minutes_actual_avg)

nine <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  select(park_date, wait_minutes_actual_avg)

wait_times <- eight |>
  left_join(nine, by = "park_date") |>
  drop_na(wait_minutes_actual_avg)
```

# Your Turn 1

First, let’s calculate the propensity score model, which will be the denominator in our stabilized weights (more to come on that soon). We’ll fit a model using `lm()` for `wait_minutes_posted_avg` with our covariates, then use the fitted predictions of `wait_minutes_posted_avg` (`.fitted`) to calculate the density using `dnorm()`.

1. Fit a model using `lm()` with `wait_minutes_posted_avg` as the outcome and the confounders identified in the DAG.
2. Use `augment()` to add model predictions to the data frame.
3. In `wt_ate()`, calculate the weights using `.fitted` and `wait_minutes_posted_avg`.

```{r}
post_time_model <- lm(
  __________,
  data = wait_times
)

wait_times_wts <- post_time_model |>
  ______(data = wait_times) |>
  mutate(
    wts = ______(
      ______,
      ______,
      exposure_type = "continuous"
    )
  )
```

# Your Turn 2

As with the example in the slides, we have a lot of extreme values for our weights

```{r}
wait_times_wts |>
  ggplot(aes(wts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") +
  scale_x_log10() +
  theme_minimal(base_size = 20) +
  xlab("Weights")
```

Let’s now fit the marginal density to use for stabilized weights:

1. Re-fit the above using stabilized weights

```{r}
wait_times_swts <- post_time_model |>
  augment(data = wait_times) |>
  mutate(
    swts = _____(
      _____,
      _____,
      exposure_type = "continuous",
      _____ = _____
    )
  )
```

Take a look at the weights now that we've stabilized them:

```{r}
ggplot(wait_times_swts, aes(swts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") +
  scale_x_log10() +
  theme_minimal(base_size = 20) +
  xlab("Stabilized Weights")
```

Stabilized weights should have a mean close to 1. If the mean is far from 1, we may have issues with model misspecification or positivity violations.

```{r}
mean(wait_times_swts$swts)
```

Here, the mean is somewhat above 1. Propensity score weights for continuous exposures are sensitive to positivity violations, and a mean like this is a signal to look closer. The bootstrap in the stretch goal is one way to do that.

# Your Turn 3

Now, let's fit the outcome model!

1. Estimate the relationship between posted wait times and actual wait times using the stabilized weights we just created, saving the model so we can pass it to `ipw()`.

```{r}
ipw_model <- lm(___ ~ ___, weights = ___, data = wait_times_swts)

ipw_model |>
  tidy() |>
  filter(term == "wait_minutes_posted_avg") |>
  mutate(estimate = estimate * 10)
```

2. Supply the propensity score model and the outcome model to `ipw()`, which uses a sandwich estimator that also accounts for the uncertainty in the propensity score model.

```{r}
ipw_fit <- ipw(____, ____)

ipw_fit
```

`ipw()` reports the estimate for one more minute of posted wait time, so multiply it by 10 to compare it to the estimate above. The point estimate matches the weighted regression; what `ipw()` adds is a standard error that accounts for having estimated the weights.

3. Switch to the conditional reading with `as_conditional()`.

```{r}
as_conditional(____) |>
  tidy(conf.int = TRUE)
```

The conditional reading reports the outcome model's own coefficients. The per-minute slope matches the marginal `slope` row here because the outcome model is linear in the exposure with an identity link, and the intercept comes with it. The marginal reading is the estimand; the conditional reading is the regression table.

## Stretch goal: Bootstrapped intervals

Bootstrap confidence intervals for our estimate.

There's nothing new here. Just remember, you need to bootstrap the entire modeling process!

```{r}
set.seed(1234)
library(rsample)

fit_model <- function(.split, ...) {
  # get bootstrapped data frame
  .df <- as.data.frame(.split)

  # fill in the rest!
}

model_estimate <- bootstraps(wait_times, 1000, apparent = TRUE) |>
  mutate(results = map(splits, ______))

# using bias-corrected confidence intervals
boot_estimate <- int_bca(_______, results, .fn = fit_model)

boot_estimate
```

--------------------------------------------------------------------------------

# Take aways

- We can calculate propensity scores for continuous exposures. `wt_ate()` uses `dnorm()` to use the normal density to transform predictions to a propensity-like scale, estimating the standard deviation from the model predictions with the pooled residual standard deviation. We can also use other approaches like quantile binning of the exposure, calculating probability-based propensity scores using categorical regression models. 
- Continuous exposures are prone to mispecification and usually need to be stabilized. A simple stabilization is to invert the propensity score by stabilization weights using an intercept-only model such as `lm(exposure ~ 1)`. `wt_ate()` can do this for you automatically with `stabilize = TRUE`. This also applies to other types of exposures.
- Stabilization is useful for any type of exposure where the weights are unbounded. Bounded weights like the ATO do not have this problem, one of their many benefits.
- Using propensity scores for continuous exposures in outcome models is identical to using them with binary exposures.
- Because propensity scores for continuous exposures are prone to positivity violation, check the bootstrap distribution of your estimate for skew and to see if the mean estimate is different from your regression model. If these problems are present, you may need to use another approach like g-computation.
